---
title: "Outcome Weights for DoubleML Objects"
subtitle: ""
author:
  - Michael Knaus
  - Roman Rakov
date: "01/26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Outcome Weights for DoubleML Objects}
  %\VignetteEngine{knitr::rmarkdown}
---

## Introduction

This notebook shows how to calculate the outcome weights ([Knaus, 2024](https://arxiv.org/abs/2411.11559)) for a [`DoubleML`](https://docs.doubleml.org/r/stable/) object. Similar to other notebooks, will use the `get_outcome_weights()` method to extract the outcome weights as described in the paper.

## **Load packages and data**

Load the relevant packages

```{r}
library(OutcomeWeights)
library(mlr3verse)
library(DoubleML)
```

## Outcome weights for DoubleML

Outcome weights can be extracted when base learners are *XGBoost*, *linear regression*, or *ranger's random forests*.

We estimate four models:

1.  Partially linear regression ([`DoubleMLPLR`](https://docs.doubleml.org/r/stable/reference/DoubleMLPLR.html)) using ranger
2.  Partially linear IV regression ([`DoubleMLPLIV`](https://docs.doubleml.org/r/stable/reference/DoubleMLPLIV.html)) using linear regression
3.  AIPW-ATE ([`DoubleMLIRM`](https://docs.doubleml.org/r/stable/reference/DoubleMLIRM.html)) using XGBoost
4.  Wald-AIPW ([`DoubleMLIIVM`](https://docs.doubleml.org/r/stable/reference/DoubleMLIIVM.html)) using XGBoost

For XGBoost, outcome weights multiplied by the outcome vector reproduce point estimates only if certain estimation practices are followed. Four hyperparameters must be set to ensure that $SY = \hat{Y}$:

```{r}
regr_xgb <- lrn("regr.xgboost", 
                   alpha = 0, 
                   subsample = 1, 
                   max_delta_step = 0, 
                   base_score = 0)

regr_lm <- lrn("regr.lm")
regr_rf <- lrn("regr.ranger", keep.inbag = TRUE)
classif_xgb <- lrn("classif.xgboost")

# PLR
ml_l_plr <- regr_rf$clone()
ml_m_plr <- regr_rf$clone()

# PLIV
ml_l_pliv <- regr_lm$clone()
ml_m_pliv <- regr_lm$clone()
ml_r_pliv <- regr_lm$clone()

# IRM (AIPW-ATE)
ml_g_irm <- regr_xgb$clone()
ml_m_irm <- classif_xgb$clone()

# IIVM (Wald-AIPW)
ml_g_iivm <- regr_xgb$clone()
ml_m_iivm <- classif_xgb$clone()
ml_r_iivm <- classif_xgb$clone()
```

Using some generic `DoubleML` data to keep it simple:

```{r}
n_obs <- 1000
dim_x <- 10
cf_folds <- 3

plr_data  <- make_plr_CCDDHNR2018(n_obs, dim_x = dim_x)
irm_data  <- make_irm_data(n_obs, dim_x = dim_x)
pliv_data <- make_pliv_CHS2015(n_obs, dim_x = dim_x, dim_z = 1)
iivm_data <- make_iivm_data(n_obs, dim_x = dim_x)

# Create DML objects
plr_obj  <- DoubleMLPLR$new(plr_data,  ml_l_plr,  ml_m_plr,  n_folds = cf_folds)
pliv_obj <- DoubleMLPLIV$new(pliv_data, ml_l_pliv, ml_m_pliv, ml_r_pliv, n_folds = cf_folds)
irm_obj  <- DoubleMLIRM$new(irm_data,  ml_g_irm, ml_m_irm, n_folds = cf_folds)
iivm_obj <- DoubleMLIIVM$new(iivm_data, ml_g_iivm, ml_m_iivm, ml_r_iivm, n_folds = cf_folds)
```

We now fit the models and extract the outcome weights that recover the `DoubleML` package output:

```{r}
plr_obj$fit(store_models = TRUE, store_predictions = TRUE)
pliv_obj$fit(store_models = TRUE, store_predictions = TRUE)
irm_obj$fit(store_models = TRUE, store_predictions = TRUE)
iivm_obj$fit(store_models = TRUE, store_predictions = TRUE)

omega_plr   <- get_outcome_weights(object = plr_obj, dml_data = plr_data)
omega_plriv <- get_outcome_weights(object = pliv_obj, dml_data = pliv_data)
omega_aipw  <- get_outcome_weights(object = irm_obj, dml_data = irm_data)
omega_waipw <- get_outcome_weights(object = iivm_obj, dml_data = iivm_data)
```

and check equivalence:

```{r}
results <- c(
  plr   = all.equal(omega_plr$omega %*% plr_data$data$y,   plr_obj$all_coef),
  pliv  = all.equal(omega_plriv$omega %*% pliv_data$data$y, pliv_obj$all_coef),
  aipw  = all.equal(omega_aipw$omega %*% irm_data$data$y,  irm_obj$all_coef),
  waipw = all.equal(omega_waipw$omega %*% iivm_data$data$y, iivm_obj$all_coef))

print(results)
```

\
