---
title: "reproduce_18"
author: "Alexander Unger"
date: "2026-04-18"
output: html_document
---


# 1. Setup

## Packages and seed

```{r, message = FALSE, warning=FALSE}
if (!require("OutcomeWeights")) install.packages("OutcomeWeights", dependencies = TRUE); library(OutcomeWeights)
if (!require("hdm"))            install.packages("hdm", dependencies = TRUE);            library(hdm)
if (!require("grf"))            install.packages("grf", dependencies = TRUE);            library(grf)
if (!require("cobalt"))         install.packages("cobalt", dependencies = TRUE);         library(cobalt)
if (!require("tidyverse"))      install.packages("tidyverse", dependencies = TRUE);      library(tidyverse)
if (!require("viridis"))        install.packages("viridis", dependencies = TRUE);        library(viridis)
if (!require("gridExtra"))      install.packages("gridExtra", dependencies = TRUE);      library(gridExtra)

set.seed(1234)
```

## Load 401(k) data

```{r}
data(pension)

D = pension$p401      # Treatment:   401(k) participation (binary)
Z = pension$e401      # Instrument:  401(k) eligibility   (binary)
Y = pension$net_tfa   # Outcome:     net total financial assets (continuous, in $)

X = model.matrix(~ 0 + age + db + educ + fsize + hown + inc + male + marr + pira + twoearn,
                 data = pension)
var_nm = c("Age","Benefit pension","Education","Family size","Home owner",
           "Income","Male","Married","IRA","Two earners")
colnames(X) = var_nm

N = length(Y)
cat("N =", N, "\n")
cat("Share treated D=1:", round(mean(D), 3), "\n")
cat("Share eligible Z=1:", round(mean(Z), 3), "\n")
cat("Share D=1 among Z=1:", round(mean(D[Z==1]), 3), "  →  no always-takers? (one-sided?)\n")
cat("Share D=1 among Z=0:", round(mean(D[Z==0]), 3), "\n")
```



---

# 2. DML Estimators (Knaus approach)

## 2a. Run Double ML with 2-fold cross-fitting

```{r}
dml_2f         = dml_with_smoother(Y, D, X, Z, n_cf_folds = 2)
results_dml_2f = summary(dml_2f)
plot(dml_2f)
```



## 2b. Extract outcome weights and verify ω'Y = θ̂

```{r}
omega_dml_2f = get_outcome_weights(dml_2f)

cat("ω'Y replicates point estimates?\n")
all.equal(as.numeric(omega_dml_2f$omega %*% Y),
          as.numeric(results_dml_2f[, 1]))

# The four DML estimators stored as rows in the omega matrix:
# Row 1: PLR       — Partially Linear Regression (no instrument)
# Row 2: PLR-IV    — Partially Linear IV (uses Z as instrument)
# Row 3: AIPW      — Augmented IPW (no instrument)
# Row 4: Wald-AIPW — Wald estimator with AIPW correction (IV, doubly robust)

w_plr      = omega_dml_2f$omega[1, ]
w_plriv    = omega_dml_2f$omega[2, ]
w_aipw     = omega_dml_2f$omega[3, ]
w_waldaipw = omega_dml_2f$omega[4, ]

theta_plriv    = sum(w_plriv * Y)
theta_waldaipw = sum(w_waldaipw * Y)

cat("\nDML point estimates (LATE candidates):\n")
cat("PLR-IV:    ", round(theta_plriv, 2), "\n")
cat("Wald-AIPW: ", round(theta_waldaipw, 2), "\n")
```



## 2c. Run Double ML with 5-fold cross-fitting

```{r}
dml_5f = dml_with_smoother(Y,D,X,Z,
                           n_cf_folds = 5)
results_dml_5f = summary(dml_5f)
plot(dml_5f)

```

---

# 3. Kappa Estimators (Słoczynski, Uysal & Wooldridge 2024)

## Background: What is Abadie's κ?

where the kappa weight for observation i is:

  κᵢ = 1  −  Dᵢ(1−Zᵢ)/(1−p(Xᵢ))  −  (1−Dᵢ)Zᵢ/p(Xᵢ)

with p(X) = P(Z=1|X) = instrument propensity score.

## 3a. Step 1 — Estimate instrument propensity score p̂(X)

```{r}
# Maximum likelihood logit — same as Słoczynski et al. "τ̂^ml" variants
ps_fit = glm(
  Z ~ age + db + educ + fsize + hown + inc + male + marr + pira + twoearn,
  data = pension,
  family = binomial(link = "logit")
)
p_hat = predict(ps_fit, type = "response")

# Basic diagnostics
summary(p_hat)
hist(p_hat, main = "Instrument propensity score p̂(X) = P(Z=1|X)",
     xlab = "p̂(X)", col = "steelblue", breaks = 40)
```



## 3b. Step 2 — Compute κ, κ₁, κ₀
```{r}
# κ
kappa = 1 - D*(1-Z)/(1-p_hat) - (1-D)*Z/p_hat

# κ1
kappa1 = D * (Z - p_hat) / (p_hat * (1 - p_hat))

# κ0
kappa0 = (1-D) * ((1-Z) - (1-p_hat)) / (p_hat * (1 - p_hat))

# correct identity
cat("κ = p̂·κ₁ + (1-p̂)·κ₀ ? Max diff:",
    max(abs(kappa - (p_hat*kappa1 + (1-p_hat)*kappa0))), "\n")

cat("mean(κ)  =", round(mean(kappa),  4), "\n")
cat("mean(κ₁) =", round(mean(kappa1), 4), "\n")
cat("mean(κ₀) =", round(mean(kappa0), 4), "\n")


# sanity check that should hold (they shoudl be more or less identical)
cat(mean(kappa), mean(kappa1), mean(kappa0))
```




## 3c. Step 3 — Build all five estimators from Słoczynski et al.

The paper studies five estimators. Here is each one with its formula:


```{r}
# Kappa estimators
tau_a =
  (sum(kappa1 * Y) / sum(kappa)) -
  (sum(kappa0 * Y) / sum(kappa))

tau_a1 =
  (sum(kappa1 * Y) / sum(kappa1)) -
  (sum(kappa0 * Y) / sum(kappa1))

tau_a0 =
  (sum(kappa1 * Y) / sum(kappa0)) -
  (sum(kappa0 * Y) / sum(kappa0))

# tau_a0 = mean(Y * (Z - p_hat) / (p_hat * (1-p_hat))) / mean(kappa0) gleiches ergebnis


tau_a10 =
  (sum(kappa1 * Y) / sum(kappa1)) -
  (sum(kappa0 * Y) / sum(kappa0))


# Numerator: IPW-weighted reduced form (effect of Z on Y)
num_z1 = sum(Z * Y / p_hat)     / sum(Z / p_hat)      # E[Y|Z=1] reweighted
num_z0 = sum((1-Z) * Y / (1-p_hat)) / sum((1-Z) / (1-p_hat))  # E[Y|Z=0] reweighted

# Denominator: IPW-weighted first stage (effect of Z on D)
den_z1 = sum(Z * D / p_hat)     / sum(Z / p_hat)      # E[D|Z=1] reweighted  
den_z0 = sum((1-Z) * D / (1-p_hat)) / sum((1-Z) / (1-p_hat))  # E[D|Z=0] reweighted

tau_u = (num_z1 - num_z0) / (den_z1 - den_z0)


w_kappa_a  <-  kappa1 / sum(kappa)  - kappa0 / sum(kappa)
w_kappa_a1 <-  kappa1 / sum(kappa1) - kappa0 / sum(kappa1)
w_kappa_a0 <-  kappa1 / sum(kappa0) - kappa0 / sum(kappa0)
w_kappa_a10 <- kappa1 / sum(kappa1) - kappa0 / sum(kappa0)

den = den_z1 - den_z0

w_kappa_u =
 (Z / p_hat) / sum(Z / p_hat) / den -
 ((1-Z) / (1-p_hat)) / sum((1-Z) / (1-p_hat)) / den


# sind die Gewichtsfaktoren konsistent mit den Schätzer : t = w*Y
all.equal(sum(w_kappa_a   * Y), tau_a)
all.equal(sum(w_kappa_a1  * Y), tau_a1)
all.equal(sum(w_kappa_a0  * Y), tau_a0)
all.equal(sum(w_kappa_a10 * Y), tau_a10)
all.equal(sum(w_kappa_u * Y), tau_u)



# Summary of all estimates

cat("\n========== LATE Estimates ==========\n")
cat("--- Kappa estimators (Słoczynski et al. 2024) ---\n")
cat("τ̂_a    (unnormalized, Abadie 2003):         ", round(tau_a,   2), "\n")
cat("τ̂_{a,1} (unnormalized, Tan/Frölich):         ", round(tau_a1,  2), "\n")
cat("τ̂_{a,0} (unnormalized, κ₀ denom):            ", round(tau_a0,  2), "\n")
cat("τ̂_{a,10} (normalized, Abadie & Cattaneo):    ", round(tau_a10, 2), "\n")
cat("τ̂_u    (normalized, Uysal 2011, RECOMMENDED):", round(tau_u,   2), "\n")
cat("\n--- DML estimators (Knaus 2024) ---\n")
cat("PLR-IV (2-fold):                             ", round(theta_plriv, 2), "\n")
cat("Wald-AIPW (2-fold):                          ", round(theta_waldaipw, 2), "\n")


```
## First Conclusion:
Der logit schätzt für Z=1 Leute einen mittleren p̂ von 0.4, also nur 46% Wahrscheinlichkeit eligible zu sein, obwohl sie es sind (Z=1). Das ist eine schlechte Schätzung. Dadurch gewichtet das IPW-Reweighting die Y-Werte der Z=1 Gruppe stark runter  und die Z=0 Gruppe stark rauf. 

```{r}
# sanity check 
if (!require("AER")) install.packages("AER", dependencies = TRUE)
library(AER)

w_u = Z / p_hat + (1 - Z) / (1 - p_hat)

iv_u_fit = AER::ivreg(Y ~ D | Z, weights = w_u)
coef(iv_u_fit)["D"]
```

---

# 4. Translation Invariance Check

This is the key diagnostic from Słoczynski et al.: unnormalized estimators are sensitive 
to adding a constant to Y (= changing units/centering), normalized ones are not.

```{r}
# Add a large constant to Y and recompute all estimators
# Under correct identification, ALL should give the same θ̂
# (because τ = E[Y1-Y0|complier] is invariant to shifting Y by a constant c)

c_shift = 10000  # shift Y by $10,000

Y_shifted = Y + c_shift

# Recompute kappa-weighted outcomes with shifted Y
tau_a_shifted =
  (sum(kappa1 * Y_shifted) / sum(kappa)) -
  (sum(kappa0 * Y_shifted) / sum(kappa))

tau_a1_shifted =
  (sum(kappa1 * Y_shifted) / sum(kappa1)) -
  (sum(kappa0 * Y_shifted) / sum(kappa1))

tau_a0_shifted =
  (sum(kappa1 * Y_shifted) / sum(kappa0)) -
  (sum(kappa0 * Y_shifted) / sum(kappa0))

tau_a10_shifted =
  (sum(kappa1 * Y_shifted) / sum(kappa1)) -
  (sum(kappa0 * Y_shifted) / sum(kappa0))

num_z1_s = sum(Z*(Y_shifted)/p_hat)/sum(Z/p_hat)
num_z0_s = sum((1-Z)*(Y_shifted)/(1-p_hat))/sum((1-Z)/(1-p_hat))
tau_u_shifted = (num_z1_s - num_z0_s) / (den_z1 - den_z0)

cat("====== Translation Invariance Test: Y → Y + 10,000 ======\n")
cat("Estimator       | Original  | Shifted   | Difference\n")
cat(sprintf("τ̂_a (unnorm.)  | %9.2f | %9.2f | %9.2f  ← NOT invariant!\n",
            tau_a, tau_a_shifted, tau_a_shifted - tau_a))
cat(sprintf("τ̂_{a,1}        | %9.2f | %9.2f | %9.2f  ← NOT invariant!\n",
            tau_a1, tau_a1_shifted, tau_a1_shifted - tau_a1))
cat(sprintf("τ̂_{a,0}        | %9.2f | %9.2f | %9.2f  ← NOT invariant!\n",
            tau_a0, tau_a0_shifted, tau_a0_shifted - tau_a0))
cat(sprintf("τ̂_{a,10}       | %9.2f | %9.2f | %9.2f  ← invariant ✓\n",
            tau_a10, tau_a10_shifted, tau_a10_shifted - tau_a10))
cat(sprintf("τ̂_u (Uysal)   | %9.2f | %9.2f | %9.2f  ← invariant ✓\n",
            tau_u, tau_u_shifted, tau_u_shifted - tau_u))
```


---

# 5. Weight Comparison: Kappa vs DML

## 5a. Distribution of weights

```{r, fig.width=12, fig.height=8}
# We compare the outcome weights of all estimators side by side

weight_list = list(
  "PLR-IV (DML)"     = w_plriv,
  "Wald-AIPW (DML)"  = w_waldaipw,
  "κ_a (unnorm.)"    = w_kappa_a,
  "κ_{a,10} (norm.)" = w_kappa_a10,
  "κ_u (Uysal, rec.)"= w_kappa_u
)

par(mfrow = c(2, 3))
for (nm in names(weight_list)) {
  w = weight_list[[nm]]
  hist(w, main = nm, xlab = "weight", col = "steelblue", breaks = 60,
       xlim = c(quantile(w, 0.005), quantile(w, 0.995)))
  abline(v = 0, col = "red", lty = 2)
}
par(mfrow = c(1,1))


weight_summary = purrr::imap_dfr(weight_list, ~tibble::tibble(
  estimator = .y,
  min = min(.x),
  p01 = quantile(.x, 0.01),
  p05 = quantile(.x, 0.05),
  median = median(.x),
  mean = mean(.x),
  p95 = quantile(.x, 0.95),
  p99 = quantile(.x, 0.99),
  max = max(.x),
  sum = sum(.x),
  sum_abs = sum(abs(.x)),
  n_negative = sum(.x < 0)
))
```



```{r}
weight_diagnostics = function(w, name) {
  data.frame(
    Estimator       = name,
    Mean            = round(mean(w), 6),
    Sum             = round(sum(w), 4),
    SD              = round(sd(w), 6),
    Min             = round(min(w), 6),
    Max             = round(max(w), 6),
    Pct_negative    = round(mean(w < 0) * 100, 2),
    Sum_abs         = round(sum(abs(w)), 3)
  )
}

diag_table = rbind(
  weight_diagnostics(w_plriv,      "PLR-IV (DML)"),
  weight_diagnostics(w_waldaipw,   "Wald-AIPW (DML)"),
  weight_diagnostics(w_kappa_a,    "κ_a (unnormalized)"),
  weight_diagnostics(w_kappa_a10,  "κ_{a,10} (normalized)"),
  weight_diagnostics(w_kappa_u,    "κ_u (Uysal, recommended)")
)

print(diag_table, row.names = FALSE)
```






# 6. Check: Covariate Balancing Propensity Score (CBPS)

Słoczynski et al. also consider covariate balancing propensity scores (CBPS) 
as an alternative to MLE logit. This often further improves balance.

```{r}
# Optional: install CBPS package for covariate balancing propensity scores
# This implements the "τ̂_u^cb" estimator — the strongest version in the paper

if (!require("CBPS")) {
  install.packages("CBPS", dependencies = TRUE)
}

if (require("CBPS")) {
  
  # Fit CBPS
  cbps_fit = CBPS::CBPS(
  Z ~ age + db + educ + fsize + hown + inc + male + marr + pira + twoearn,
  data = pension,
  ATT = 0,
  method = "exact",
  standardize = FALSE
)
  p_hat_cb = fitted(cbps_fit)
  
  cat("CBPS propensity score summary:\n")
  print(summary(p_hat_cb))
  
  # Recompute κ₁ and the Uysal estimator with CBPS propensity scores
  kappa1_cb = D * (Z/p_hat_cb - (1-Z)/(1-p_hat_cb))
  kappa0_cb = (1-D) * ((1-Z)/(1-p_hat_cb) - Z/p_hat_cb)
  
  num_z1_cb = sum(Z*Y/p_hat_cb)/sum(Z/p_hat_cb)
  num_z0_cb = sum((1-Z)*Y/(1-p_hat_cb))/sum((1-Z)/(1-p_hat_cb))
  den_z1_cb = sum(Z*D/p_hat_cb)/sum(Z/p_hat_cb)
  den_z0_cb = sum((1-Z)*D/(1-p_hat_cb))/sum((1-Z)/(1-p_hat_cb))
  
  tau_u_cb = (num_z1_cb - num_z0_cb) / (den_z1_cb - den_z0_cb)
  
  cat("\nτ̂_u with MLE propensity score:", round(tau_u, 2), "\n")
  cat("τ̂_u with CBPS propensity score:", round(tau_u_cb, 2), "\n")
  
} else {
  cat("CBPS package not available. Skip this section.\n")
}
hist(p_hat_cb, main = "Instrument propensity score p̂(X) estimated with CBPS ",
     xlab = "p̂(X)", col = "steelblue", breaks = 40)
```



```{r}
kappa_cb  = 1 - D*(1-Z)/(1-p_hat_cb) - (1-D)*Z/p_hat_cb
kappa1_cb = D * (Z/p_hat_cb - (1-Z)/(1-p_hat_cb))
kappa0_cb = (1-D) * ((1-Z)/(1-p_hat_cb) - Z/p_hat_cb)

tau_a1_cb =
  (sum(kappa1_cb * Y) / sum(kappa1_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa1_cb))

tau_a0_cb =
  (sum(kappa1_cb * Y) / sum(kappa0_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa0_cb))

tau_a10_cb =
  (sum(kappa1_cb * Y) / sum(kappa1_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa0_cb))


c(
  tau_u_cb   = tau_u_cb,
  tau_a1_cb  = tau_a1_cb,
  tau_a0_cb  = tau_a0_cb,
  tau_a10_cb = tau_a10_cb
)


cat("tau_u^ml   =", round(tau_u, 2), "\n")
cat("tau_u^cb   =", round(tau_u_cb, 2), "\n")
cat("tau_a1^cb  =", round(tau_a1_cb, 2), "\n")
cat("tau_a0^cb  =", round(tau_a0_cb, 2), "\n")
cat("tau_a10^cb =", round(tau_a10_cb, 2), "\n")
cat("mean(kappa1_cb) =", round(mean(kappa1_cb), 6), "\n")
cat("mean(kappa0_cb) =", round(mean(kappa0_cb), 6), "\n")





```



# 6b CBPS via Newton-Raphson (Imai & Ratkovic 2014)
```{r}
# ============================================================
# CBPS via Newton-Raphson (Imai & Ratkovic 2014)
# Löst die Momentbedingung: E[Z/p(X) * X] = E[(1-Z)/(1-p(X)) * X]
# Ersatz für das CBPS-Paket mit ATT=0
# ============================================================
cbps_iv <- function(Z, X_mat, tol = 1e-8, max_iter = 1000) {
  
  sigmoid <- function(x) 1 / (1 + exp(-x))
  
  # Momentbedingung: soll = 0 sein
  moment_fn <- function(b) {
    p <- pmin(pmax(sigmoid(X_mat %*% b), 1e-10), 1 - 1e-10)
    colMeans(as.vector(Z / p - (1 - Z) / (1 - p)) * X_mat)
  }
  
  # Jacobian der Momentbedingung
  jac_fn <- function(b) {
    p <- pmin(pmax(sigmoid(X_mat %*% b), 1e-10), 1 - 1e-10)
    w <- as.vector(-Z * (1 - p) / p - (1 - Z) * p / (1 - p))
    t(w * X_mat) %*% X_mat / length(Z)
  }
  
  # Warmstart: MLE-Logit
  b <- coef(glm(Z ~ X_mat - 1, family = binomial(link = "logit")))
  
  for (i in seq_len(max_iter)) {
    m <- moment_fn(b)
    if (max(abs(m)) < tol) break
    
    step <- tryCatch(solve(jac_fn(b), -m), error = function(e) NULL)
    if (is.null(step)) break
    
    # Backtracking line search
    alpha <- 1.0
    for (j in seq_len(30)) {
      if (max(abs(moment_fn(b + alpha * step))) < max(abs(m))) break
      alpha <- alpha * 0.5
    }
    b <- b + alpha * step
  }
  
  p <- pmin(pmax(sigmoid(X_mat %*% b), 1e-10), 1 - 1e-10)
  list(
    p_hat     = as.vector(p),
    converged = max(abs(moment_fn(b))) < 1e-5
  )
}



# NEU — einsetzen:
X_mat <- model.matrix(~ age + db + educ + fsize + hown +
                        inc + male + marr + pira + twoearn,
                      data = pension)

cb        <- cbps_iv(Z, X_mat)
p_hat_cb  <- cb$p_hat
cat("CBPS converged:", cb$converged, "\n")


kappa_cb  = 1 - D*(1-Z)/(1-p_hat_cb) - (1-D)*Z/p_hat_cb
kappa1_cb = D * (Z - p_hat_cb) / (p_hat_cb * (1 - p_hat_cb))
kappa0_cb = (1-D) * ((1-Z) - (1 - p_hat_cb)) / (p_hat_cb * (1 - p_hat_cb))

tau_a1_cb =
  (sum(kappa1_cb * Y) / sum(kappa1_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa1_cb))

tau_a0_cb =
  (sum(kappa1_cb * Y) / sum(kappa0_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa0_cb))

tau_a10_cb =
  (sum(kappa1_cb * Y) / sum(kappa1_cb)) -
  (sum(kappa0_cb * Y) / sum(kappa0_cb))

num_z1_cb <- sum(Z * Y / p_hat_cb) / sum(Z / p_hat_cb)
num_z0_cb <- sum((1-Z) * Y / (1-p_hat_cb)) / sum((1-Z) / (1-p_hat_cb))
den_z1_cb <- sum(Z * D / p_hat_cb) / sum(Z / p_hat_cb)
den_z0_cb <- sum((1-Z) * D / (1-p_hat_cb)) / sum((1-Z) / (1-p_hat_cb))
tau_u_cb  <- (num_z1_cb - num_z0_cb) / (den_z1_cb - den_z0_cb)


c(
  tau_u_cb   = tau_u_cb,
  tau_a1_cb  = tau_a1_cb,
  tau_a0_cb  = tau_a0_cb,
  tau_a10_cb = tau_a10_cb
)



cat("tau_u^ml   =", round(tau_u, 2), "\n")
cat("tau_u^cb   =", round(tau_u_cb, 2), "\n")
cat("tau_a1^cb  =", round(tau_a1_cb, 2), "\n")
cat("tau_a0^cb  =", round(tau_a0_cb, 2), "\n")
cat("tau_a10^cb =", round(tau_a10_cb, 2), "\n")
cat("mean(kappa1_cb) =", round(mean(kappa1_cb), 6), "\n")
cat("mean(kappa0_cb) =", round(mean(kappa0_cb), 6), "\n")



```







# 7. Check: Trying to debug: The naive Wald
```{r}
# Vergleich naive LATE ohne Gewichte
late_naive = mean(Y[Z==1]) - mean(Y[Z==0])
fs_naive   = mean(D[Z==1]) - mean(D[Z==0])
cat("\nNaiver Wald-Schätzer (ohne X):\n")
cat("Reduced form:", round(late_naive, 2), "\n")
cat("First stage: ", round(fs_naive, 4), "\n")
cat("LATE = RF/FS:", round(late_naive/fs_naive, 2), "\n")




```



