---
title: "Card (1995) — Design Diagnostics & Weight Stability"
subtitle: "Why is the Effective Sample Size so Low? Propensity Scores, Overlap, and Weak Instrument"
author: "Alexander Unger"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
    df_print: kable
---

```{css, echo=FALSE}
body { font-size: 15px; color: #2c2c2c; }
h1   { border-bottom: 2px solid #2c7bb6; padding-bottom: 6px; margin-top: 36px; }
h2   { color: #2c7bb6; margin-top: 28px; }
h3   { color: #444; margin-top: 20px; }
.note-box { background:#eaf4ff; border-left:4px solid #2c7bb6; padding:10px 16px; border-radius:4px; margin:14px 0; font-size:14px; }
.warn-box { background:#fff3e0; border-left:4px solid #fb8c00; padding:10px 16px; border-radius:4px; margin:14px 0; font-size:14px; }
.key-box  { background:#fdeaea; border-left:4px solid #d7191c; padding:10px 16px; border-radius:4px; margin:14px 0; font-size:14px; }
table { font-size: 13px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 5.5, dpi = 110)
```

```{r packages}
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(AER)
library(sandwich)
library(lmtest)
library(scales)

# Shared kappa toolkit (logit_mle, get_cbps_p, kappa_outcome_weights,
# kappa_weights, weight_diag, ...). Nothing estimation-related is defined here.
source("functions_all.R")

# Consistent palette
col_z1   <- "#2c7bb6"   # Z = 1 (near college)
col_z0   <- "#d7191c"   # Z = 0 (far from college)
col_grey <- "#4d4d4d"
col_card <- "#2166ac"
col_kit  <- "#984ea3"
```

```{r data}
data_path <- "/Users/alexung/Desktop/master_thesis/code/"
card      <- read_dta(file.path(data_path, "card.dta"))
card      <- card[!is.na(card$wage) & !is.na(card$educ), ]

# Treatments, instrument, covariates
card$somecol <- as.integer(card$educ > 12)    # D1: some college
card$educ16  <- as.integer(card$educ >= 16)   # D2: college completion
card$Z       <- as.integer(card$nearc4)       # Z: grew up near 4-yr college
if (!"expersq" %in% names(card)) card$expersq <- card$exper^2

# Propensity scores p(X) = Pr(Z = 1 | X) under both specs
X_card <- model.matrix(~ exper + expersq + black + south + smsa + smsa66 +
                          reg661+reg662+reg663+reg664+reg665+reg666+reg667+reg668, card)
X_kit  <- model.matrix(~ black + south + smsa + smsa66 + south66, card)

card$p_card <- logit_mle(card$Z, X_card)
card$p_kit  <- logit_mle(card$Z, X_kit)

cat(sprintf("N = %d | Pr(Z=1) = %.3f | somecol = %.3f | educ16 = %.3f\n",
            nrow(card), mean(card$Z), mean(card$somecol), mean(card$educ16)))
```

---

# 1. The Puzzle

<div class="key-box">
**The finding to explain.** In the Card application the effective sample size (ESS = $1/\sum_i \omega_i^2$) of the kappa weights collapses to single digits — for the Kitagawa specification it falls near 1 — even though $N = 3{,}010$. This notebook diagnoses *why*, working through three candidate explanations: (1) a **weak first stage**, (2) **propensity-score concentration / non-overlap**, and (3) **sign-reversing observations** (high $\hat p(X)$ yet $Z=0$ and treated, or the reverse) that destabilize the kappa weights.
</div>

```{r ess-preview}
# Quick ESS preview across both specs and both treatments
ess_one <- function(D, p) {
  w <- kappa_outcome_weights(card$Z, D, p)$w_u
  1 / sum(w^2)
}
ess_tab <- data.frame(
  Specification = c("Card", "Kitagawa"),
  `somecol (ESS)` = c(ess_one(card$somecol, card$p_card),
                      ess_one(card$somecol, card$p_kit)),
  `educ16 (ESS)`  = c(ess_one(card$educ16,  card$p_card),
                      ess_one(card$educ16,  card$p_kit)),
  check.names = FALSE
)
ess_tab[,2:3] <- round(ess_tab[,2:3], 1)

ess_tab %>%
  kbl(caption = "Effective sample size (τ_ml_u weights) — the puzzle in one table",
      align = c("l","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE, font_size = 14) %>%
  column_spec(2:3, bold = TRUE, color = "#d7191c") %>%
  footnote(general = sprintf("Out of N = %d observations. ESS this low means a handful of observations carry essentially all the weight.", nrow(card)))
```

---

# 2. First Stage: Is the Instrument Weak?

<div class="note-box">
**Why this matters.** The kappa weights divide by the conditional first stage. A weak instrument means $\Pr(D=1\mid Z=1,X) - \Pr(D=1\mid Z=0,X)$ is small for many cells; dividing by a near-zero first stage inflates weights and destroys ESS. Card's instrument (college proximity) is known to be weaker than the Vietnam draft lottery.
</div>

```{r first-stage-table}
fs_F <- function(formula, data) {
  fit <- lm(formula, data = data)
  ct  <- coeftest(fit, vcov = vcovHC(fit, type = "HC1"))
  c(coef = unname(ct["Z","Estimate"]),
    se   = unname(ct["Z","Std. Error"]),
    F    = unname((ct["Z","Estimate"]/ct["Z","Std. Error"])^2))
}

# First stage for both treatments and both specs
fs_rows <- rbind(
  data.frame(Treatment="somecol", Spec="Card",
             t(fs_F(somecol ~ Z + exper+expersq+black+south+smsa+smsa66+
                      reg661+reg662+reg663+reg664+reg665+reg666+reg667+reg668, card))),
  data.frame(Treatment="somecol", Spec="Kitagawa",
             t(fs_F(somecol ~ Z + black+south+smsa+smsa66+south66, card))),
  data.frame(Treatment="educ16", Spec="Card",
             t(fs_F(educ16 ~ Z + exper+expersq+black+south+smsa+smsa66+
                      reg661+reg662+reg663+reg664+reg665+reg666+reg667+reg668, card))),
  data.frame(Treatment="educ16", Spec="Kitagawa",
             t(fs_F(educ16 ~ Z + black+south+smsa+smsa66+south66, card)))
)
fs_rows$coef <- round(fs_rows$coef, 4)
fs_rows$se   <- round(fs_rows$se, 4)
fs_rows$F    <- round(fs_rows$F, 1)
fs_rows$`Stock-Yogo (>10)?` <- ifelse(fs_rows$F >= 10, "✓ strong", "✗ WEAK")

fs_rows %>%
  kbl(caption = "First-stage strength: effect of college proximity (Z) on each treatment",
      col.names = c("Treatment","Spec","First stage δ̂","SE","F-stat","Stock-Yogo (>10)?"),
      align = c("l","l","r","r","r","c")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 13) %>%
  row_spec(which(fs_rows$F < 10), background = "#fdeaea", bold = TRUE) %>%
  footnote(general = "HC1-robust first-stage F on the instrument. Compare to Vietnam, where F was far above 10. A weak first stage here is the first driver of weight instability.")
```

```{r first-stage-plot}
# First-stage gap: D-rate by Z, for somecol
fs_gap <- card %>%
  summarise(
    `Z = 1 (near college)` = mean(somecol[Z == 1]),
    `Z = 0 (far)`          = mean(somecol[Z == 0])
  )
gap_some <- fs_gap$`Z = 1 (near college)` - fs_gap$`Z = 0 (far)`

fs_gap_e <- card %>%
  summarise(
    `Z = 1 (near college)` = mean(educ16[Z == 1]),
    `Z = 0 (far)`          = mean(educ16[Z == 0])
  )
gap_educ <- fs_gap_e$`Z = 1 (near college)` - fs_gap_e$`Z = 0 (far)`

fs_plot_df <- data.frame(
  Treatment = rep(c("somecol\n(educ > 12)", "educ16\n(educ ≥ 16)"), each = 2),
  Z = rep(c("Z = 1 (near college)", "Z = 0 (far)"), 2),
  rate = c(fs_gap$`Z = 1 (near college)`, fs_gap$`Z = 0 (far)`,
           fs_gap_e$`Z = 1 (near college)`, fs_gap_e$`Z = 0 (far)`)
)

ggplot(fs_plot_df, aes(x = Treatment, y = rate, fill = Z)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.9) +
  geom_text(aes(label = sprintf("%.1f%%", 100*rate)),
            position = position_dodge(0.7), vjust = -0.4, size = 3.4) +
  annotate("text", x = 1, y = 0.55,
           label = sprintf("gap = %.3f", gap_educ), size = 3.5, colour = col_grey) +
  annotate("text", x = 2, y = 0.55,
           label = sprintf("gap = %.3f", gap_some), size = 3.5, colour = col_grey) +
  scale_fill_manual(values = c("Z = 1 (near college)" = col_z1, "Z = 0 (far)" = col_z0)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.62)) +
  labs(title = "First stage: treatment rates by college proximity",
       subtitle = "The gap between bars is the unconditional first stage. Small gaps = weak instrument.",
       x = NULL, y = "Treatment rate", fill = NULL,
       caption = "Card (1995), N = 3,010.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 10),
        plot.caption = element_text(colour = "grey50", size = 9))
```

---

# 3. Propensity Score: Concentration & Overlap

<div class="note-box">
**Why this matters.** The kappa weights contain $1/p(X)$ and $1/(1-p(X))$ terms. Observations with $\hat p(X)$ near 0 or 1 get enormous weights. Unlike Vietnam (where $\hat p$ was tightly clustered around 0.46 by design), the richer Card and Kitagawa covariate sets can push $\hat p(X)$ toward the boundaries — directly inflating weights.
</div>

## Summary Statistics

```{r ps-summary}
ps_summary <- function(p, name) {
  data.frame(Specification = name,
             Min = min(p), P05 = quantile(p,0.05), P25 = quantile(p,0.25),
             Mean = mean(p), P75 = quantile(p,0.75), P95 = quantile(p,0.95),
             Max = max(p), SD = sd(p),
             `near 0/1` = mean(p < 0.05 | p > 0.95),
             check.names = FALSE)
}
ps_tab <- bind_rows(ps_summary(card$p_card, "Card (14 covariates)"),
                    ps_summary(card$p_kit,  "Kitagawa (5 covariates)"))
ps_tab[,2:10] <- lapply(ps_tab[,2:10], function(x) round(x, 4))

ps_tab %>%
  kbl(caption = "Propensity score p̂(X) = Pr(Z=1|X): summary by specification",
      align = c("l", rep("r", 9))) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 12.5) %>%
  column_spec(9, bold = TRUE) %>%
  column_spec(10, bold = TRUE, color = "#d7191c") %>%
  footnote(general = "Compare SD and the 'near 0/1' share against Vietnam (SD ≈ 0.10, essentially nothing near the boundaries). Higher dispersion here is the second driver of low ESS.")
```

## Distribution by Instrument Status

```{r ps-dist-plot, fig.height=6}
ps_long <- card %>%
  select(p_card, p_kit, Z) %>%
  pivot_longer(c(p_card, p_kit), names_to = "spec", values_to = "p_hat") %>%
  mutate(Specification = factor(spec,
                                levels = c("p_card", "p_kit"),
                                labels = c("Card spec", "Kitagawa spec")),
         Zlab = factor(ifelse(Z == 1, "Z = 1 (near college)", "Z = 0 (far)"),
                       levels = c("Z = 1 (near college)", "Z = 0 (far)")))

ggplot(ps_long, aes(x = p_hat, fill = Zlab)) +
  geom_histogram(bins = 40, alpha = 0.7, colour = "white", position = "identity") +
  geom_vline(xintercept = mean(card$Z), linetype = "dashed", colour = col_grey) +
  facet_wrap(~ Specification, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("Z = 1 (near college)" = col_z1, "Z = 0 (far)" = col_z0)) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Propensity score distribution by instrument status",
       subtitle = "Overlap requires both colours present at every p̂. Gaps at the tails = overlap violations → exploding weights.",
       x = expression(hat(p)(X) == Pr(Z == 1 ~ "|" ~ X)), y = "Count", fill = NULL,
       caption = "Dashed line = pooled Pr(Z=1).") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9.5),
        plot.caption = element_text(colour = "grey50", size = 9))
```

<div class="warn-box">
**What to look for.** Where the red (Z=0) bars vanish at high $\hat p$, or blue (Z=1) bars vanish at low $\hat p$, the inverse-probability terms $Z/p$ and $(1-Z)/(1-p)$ have no counterpart to balance against — those observations receive extreme weights. The Kitagawa panel, with only 5 binary covariates, produces discrete $\hat p$ clumps that are especially prone to this.
</div>

---

# 4. Where Do the Weights Explode?

<div class="note-box">
**The mechanism.** The kappa weight for the normalized estimator is built from $Z/p - (1-Z)/(1-p)$ scaled by the first stage. Plotting each observation's absolute weight against its $\hat p(X)$ shows exactly which observations dominate — and confirms whether it is the tails of $\hat p$, the sign-reversing cells, or both.
</div>

```{r weight-vs-pscore, fig.height=6}
# Build per-observation weights for the worst cell (Kitagawa, the low-ESS one)
make_wdf <- function(D, p, spec, treat) {
  w <- kappa_outcome_weights(card$Z, D, p)$w_u
  data.frame(p_hat = p, w = w, abs_w = abs(w),
             Z = card$Z, D = D,
             Spec = spec, Treat = treat)
}
wdf <- bind_rows(
  make_wdf(card$somecol, card$p_kit,  "Kitagawa", "somecol"),
  make_wdf(card$somecol, card$p_card, "Card",     "somecol")
)
wdf$cell <- paste(wdf$Spec, "spec")
wdf$status <- factor(
  ifelse(wdf$Z == 1 & wdf$D == 1, "Z=1, treated",
  ifelse(wdf$Z == 1 & wdf$D == 0, "Z=1, untreated",
  ifelse(wdf$Z == 0 & wdf$D == 1, "Z=0, treated",
                                  "Z=0, untreated"))),
  levels = c("Z=1, treated","Z=1, untreated","Z=0, treated","Z=0, untreated"))

ggplot(wdf, aes(x = p_hat, y = abs_w, colour = status)) +
  geom_point(alpha = 0.45, size = 1.1) +
  facet_wrap(~ cell, ncol = 2, scales = "free") +
  scale_colour_manual(values = c("Z=1, treated"   = "#2c7bb6",
                                 "Z=1, untreated" = "#abd9e9",
                                 "Z=0, treated"   = "#d7191c",
                                 "Z=0, untreated" = "#fdae61")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Absolute outcome weight vs. propensity score (somecol)",
       subtitle = "Each point is one observation. Points high on the y-axis dominate the estimate and collapse the ESS.",
       x = expression(hat(p)(X)), y = expression("|" ~ omega[i] ~ "|"),
       colour = NULL,
       caption = "Watch for: 'Z=0, treated' at high p̂ and 'Z=1, untreated' at low p̂ — these sign-reversing cells generate the largest weights.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40", size = 9.5),
        plot.caption = element_text(colour = "grey50", size = 8.8))
```

## The Top-Weighted Observations

```{r top-weights}
# How concentrated? Show how much total weight the top-k observations carry.
conc_one <- function(D, p, spec) {
  w  <- kappa_outcome_weights(card$Z, D, p)$w_u
  aw <- sort(abs(w), decreasing = TRUE)
  tot <- sum(aw)
  data.frame(Spec = spec,
             `Top 1 (%)`  = round(100 * aw[1] / tot, 1),
             `Top 5 (%)`  = round(100 * sum(aw[1:5]) / tot, 1),
             `Top 10 (%)` = round(100 * sum(aw[1:10]) / tot, 1),
             `Top 25 (%)` = round(100 * sum(aw[1:25]) / tot, 1),
             ESS = round(1 / sum(w^2), 1),
             check.names = FALSE)
}
conc_tab <- bind_rows(
  conc_one(card$somecol, card$p_card, "Card / somecol"),
  conc_one(card$somecol, card$p_kit,  "Kitagawa / somecol"),
  conc_one(card$educ16,  card$p_card, "Card / educ16"),
  conc_one(card$educ16,  card$p_kit,  "Kitagawa / educ16")
)

conc_tab %>%
  kbl(caption = "Weight concentration: share of total |ω| carried by the most extreme observations",
      align = c("l","r","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(6, bold = TRUE, color = "#d7191c") %>%
  footnote(general = "If a handful of observations carry most of the weight, the estimator is effectively using only those points — that is what the low ESS reports numerically.")
```

---

# 5. Overlap Diagnosis by Treatment Cell

<div class="note-box">
**Putting it together.** The cross-tabulation of $Z$ × $D$ within propensity-score bins shows where compliers actually live and where cells are empty or near-empty. Empty diagonal cells (e.g. high-$\hat p$ but no $Z=0$ controls) are the formal overlap violations.
</div>

```{r overlap-table}
overlap_by_bin <- function(D, p, spec) {
  bin <- cut(p, breaks = quantile(p, seq(0,1,0.2)), include.lowest = TRUE,
             labels = c("Q1 (low p̂)","Q2","Q3","Q4","Q5 (high p̂)"))
  out <- card %>%
    mutate(bin = bin, D = D) %>%
    group_by(bin) %>%
    summarise(n = n(),
              `n(Z=1)` = sum(Z == 1),
              `n(Z=0)` = sum(Z == 0),
              `Pr(D|Z=1)` = round(mean(D[Z==1]), 3),
              `Pr(D|Z=0)` = round(ifelse(sum(Z==0)>0, mean(D[Z==0]), NA), 3),
              `first stage` = round(mean(D[Z==1]) - ifelse(sum(Z==0)>0, mean(D[Z==0]), NA), 3),
              .groups = "drop")
  out$Spec <- spec
  out
}
ov_kit <- overlap_by_bin(card$somecol, card$p_kit, "Kitagawa")

ov_kit %>%
  select(bin, n, `n(Z=1)`, `n(Z=0)`, `Pr(D|Z=1)`, `Pr(D|Z=0)`, `first stage`) %>%
  kbl(caption = "Overlap by propensity-score quintile — Kitagawa spec, somecol (the low-ESS cell)",
      align = c("l","r","r","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 13) %>%
  row_spec(which(ov_kit$`n(Z=0)` < 30), background = "#fdeaea") %>%
  footnote(general = "Highlighted rows: bins with very few Z=0 observations — thin overlap. A near-zero or negative 'first stage' within a bin is where the kappa weights flip sign and explode.")
```

---

# 6. Summary: Three Drivers of the Low ESS

<div class="key-box">
**For the meeting.** The low ESS in Card is not an error — it is the visible symptom of three reinforcing design features, all absent in the Vietnam application:

1. **Weak first stage** (Section 2). College proximity moves treatment far less than the draft lottery moved veteran status, so the conditional first stage in the weight denominator is small for many cells.

2. **Dispersed / boundary propensity scores** (Section 3). The richer Card and Kitagawa covariate sets push $\hat p(X)$ toward 0 and 1, where the $1/p$ and $1/(1-p)$ terms blow up — unlike Vietnam's tightly clustered $\hat p \approx 0.46$.

3. **Sign-reversing cells & thin overlap** (Sections 4–5). A small number of observations with high $\hat p$ but $Z=0$ (or low $\hat p$ but $Z=1$) receive enormous weights and carry most of the total — the ESS simply counts how few observations effectively drive the estimate.

The Kitagawa specification is worst on all three counts, which is exactly why its ESS approaches 1.
</div>

```{r vietnam-contrast}
contrast <- data.frame(
  Feature = c("First-stage F (typical)",
              "p̂(X) standard deviation",
              "Share of p̂ near 0 or 1",
              "Resulting ESS"),
  Vietnam = c("≫ 10 (strong)", "≈ 0.10 (tight)", "≈ 0%", "uniform, moderate"),
  `Card / Kitagawa` = c("can fall below 10", 
                        sprintf("%.3f / %.3f", sd(card$p_card), sd(card$p_kit)),
                        sprintf("%.1f%% / %.1f%%",
                                100*mean(card$p_card<0.05 | card$p_card>0.95),
                                100*mean(card$p_kit<0.05  | card$p_kit>0.95)),
                        "collapses toward 1"),
  check.names = FALSE
)
contrast %>%
  kbl(caption = "Why Card differs from Vietnam — side by side",
      align = c("l","l","l")) %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 13) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)
```

---

*Design diagnostics for the Card (1995) application.*  
*All weight/propensity functions sourced from `functions_all.R`.*
