---
title: "child_descr_30"
author: "Alexander Unger"
date: "2026-05-30"
output: html_document
---

```{r}
# ==============================================================================
# DESCRIPTIVE BLOCK 1 — Setup, data loading, variable construction, sample overview
# ==============================================================================
# Angrist & Evans (1998) childbearing design, following SUW Table 4.
#
# Design:
#   Z = samesex   : 1 if first two children have the same sex
#   D = morekids  : 1 if mother has at least three children
#
# Outcomes:
#   LFP sample:
#     workedm      = 1 if worked for pay, 0 otherwise
#     workedm_21   = workedm + 1
#     workedm_01   = 1 - workedm
#
#   Positive-income sample:
#     lincomem      = log income in dollars
#     lincomem_cnt  = log income in cents
#     lincomem_ths  = log income in thousands
#     lincomem_hths = log income in hundred-thousands
#
# Covariates:
#   agem, agefstm, sexk, sex2nd, blackm, hispm, othracem
# ==============================================================================


# ------------------------------------------------------------------------------
# 0. Packages
# ------------------------------------------------------------------------------

required_packages <- c(
  "haven",
  "knitr"
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(
      sprintf(
        "Package '%s' is required but not installed. Install it with install.packages('%s').",
        pkg, pkg
      )
    )
  }
}

library(haven)
library(knitr)


# ------------------------------------------------------------------------------
# 1. User path
# ------------------------------------------------------------------------------
# Adjust this path if ae98.dta is stored somewhere else.

data_path <- "/Users/alexung/Desktop/master_thesis/code/"
ae98_file <- file.path(data_path, "ae98.dta")

if (!file.exists(ae98_file)) {
  stop(
    paste0(
      "Could not find ae98.dta at:\n",
      ae98_file,
      "\n\nPlease change data_path at the top of this block."
    )
  )
}


# ------------------------------------------------------------------------------
# 2. Helper functions for readable output
# ------------------------------------------------------------------------------

show_desc_table <- function(x, caption = NULL, digits = 3) {
  x_print <- as.data.frame(x)

  num_cols <- sapply(x_print, is.numeric)
  x_print[num_cols] <- lapply(x_print[num_cols], function(v) round(v, digits))

  if (isTRUE(getOption("knitr.in.progress"))) {
    knitr::kable(
      x_print,
      format = "html",
      row.names = FALSE,
      caption = caption
    )
  } else {
    if (!is.null(caption)) {
      cat("\n", caption, "\n", sep = "")
    }
    print(x_print, row.names = FALSE)
    invisible(x_print)
  }
}

binary_summary <- function(x) {
  x <- as.numeric(x)

  c(
    N = sum(!is.na(x)),
    Mean = mean(x, na.rm = TRUE),
    Share_0 = mean(x == 0, na.rm = TRUE),
    Share_1 = mean(x == 1, na.rm = TRUE),
    Missing = sum(is.na(x))
  )
}

continuous_summary <- function(x) {
  x <- as.numeric(x)

  c(
    N = sum(!is.na(x)),
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Q25 = unname(quantile(x, 0.25, na.rm = TRUE)),
    Median = median(x, na.rm = TRUE),
    Q75 = unname(quantile(x, 0.75, na.rm = TRUE)),
    Max = max(x, na.rm = TRUE),
    Missing = sum(is.na(x))
  )
}


# ------------------------------------------------------------------------------
# 3. Read raw data
# ------------------------------------------------------------------------------

ae98_raw <- haven::read_dta(ae98_file)

cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 1 — Setup and sample construction\n")
cat("============================================================\n\n")

cat("Raw AE98 data loaded from:\n")
cat("  ", ae98_file, "\n\n", sep = "")

cat("Raw number of observations:", nrow(ae98_raw), "\n")
cat("Raw number of variables:   ", ncol(ae98_raw), "\n\n")


# ------------------------------------------------------------------------------
# 4. Check required variables exist
# ------------------------------------------------------------------------------

required_vars <- c(
  "workedm",
  "morekids",
  "samesex",
  "incomem",
  "agem",
  "agefstm",
  "sexk",
  "sex2nd",
  "blackm",
  "hispm",
  "othracem"
)

missing_vars <- setdiff(required_vars, names(ae98_raw))

if (length(missing_vars) > 0) {
  stop(
    paste0(
      "The following required variables are missing from ae98.dta:\n",
      paste(missing_vars, collapse = ", ")
    )
  )
}

cat("All required variables found:\n")
cat("  ", paste(required_vars, collapse = ", "), "\n\n", sep = "")


# ------------------------------------------------------------------------------
# 5. Main sample restriction
# ------------------------------------------------------------------------------
# We follow the replication logic:
#   - LFP requires workedm, morekids, and samesex.
#   - Income additionally requires incomem > 0 because log income is undefined
#     for zero or negative income.

ae98 <- ae98_raw[
  !is.na(ae98_raw$workedm) &
    !is.na(ae98_raw$morekids) &
    !is.na(ae98_raw$samesex),
]

cat("After main LFP restriction:\n")
cat("  N =", nrow(ae98), "\n\n")


# ------------------------------------------------------------------------------
# 6. Outcome construction
# ------------------------------------------------------------------------------

# Labor-force-participation recodings
ae98$workedm_21 <- ae98$workedm + 1
ae98$workedm_01 <- 1 - ae98$workedm

# Log-income unit transformations
# Note:
#   lincomem variables are NA/Inf for non-positive income if computed directly.
#   We construct them for the full object, but only use ae98_income for analysis.
ae98$lincomem      <- ifelse(ae98$incomem > 0, log(ae98$incomem), NA_real_)
ae98$lincomem_cnt  <- ifelse(ae98$incomem > 0, log(100 * ae98$incomem), NA_real_)
ae98$lincomem_ths  <- ifelse(ae98$incomem > 0, log(ae98$incomem / 1000), NA_real_)
ae98$lincomem_hths <- ifelse(ae98$incomem > 0, log(ae98$incomem / 100000), NA_real_)


# ------------------------------------------------------------------------------
# 7. Define analysis samples
# ------------------------------------------------------------------------------

ae98_lfp <- ae98

ae98_income <- ae98[
  !is.na(ae98$incomem) &
    ae98$incomem > 0 &
    !is.na(ae98$lincomem),
]

cat("Analysis samples:\n")
cat("  N (LFP sample, ae98_lfp):          ", nrow(ae98_lfp), "\n", sep = "")
cat("  N (income sample, ae98_income):    ", nrow(ae98_income), "\n", sep = "")
cat("  Share retained for income sample:  ",
    round(nrow(ae98_income) / nrow(ae98_lfp), 4), "\n\n", sep = "")


# ------------------------------------------------------------------------------
# 8. Covariate matrix construction
# ------------------------------------------------------------------------------

make_X_ae <- function(df) {
  model.matrix(
    ~ agem + agefstm + sexk + sex2nd + blackm + hispm + othracem,
    data = df
  )
}

X_lfp <- make_X_ae(ae98_lfp)
X_income <- make_X_ae(ae98_income)

cat("Covariate matrices:\n")
cat("  dim(X_lfp):    ", paste(dim(X_lfp), collapse = " x "), "\n", sep = "")
cat("  dim(X_income): ", paste(dim(X_income), collapse = " x "), "\n", sep = "")
cat("  X columns:     ", paste(colnames(X_lfp), collapse = ", "), "\n\n", sep = "")


# ------------------------------------------------------------------------------
# 9. Sample construction overview table
# ------------------------------------------------------------------------------

sample_overview <- data.frame(
  Sample = c(
    "Raw AE98 data",
    "LFP sample",
    "Positive-income sample"
  ),
  Object = c(
    "ae98_raw",
    "ae98_lfp",
    "ae98_income"
  ),
  Restriction = c(
    "As loaded from ae98.dta",
    "Non-missing workedm, morekids, and samesex",
    "LFP sample restricted to incomem > 0"
  ),
  N = c(
    nrow(ae98_raw),
    nrow(ae98_lfp),
    nrow(ae98_income)
  ),
  Share_of_LFP_sample = c(
    NA_real_,
    1,
    nrow(ae98_income) / nrow(ae98_lfp)
  ),
  stringsAsFactors = FALSE
)

show_desc_table(
  sample_overview,
  caption = "Sample construction overview"
)


# ------------------------------------------------------------------------------
# 10. Core design variables in both samples
# ------------------------------------------------------------------------------

core_lfp <- rbind(
  samesex  = binary_summary(ae98_lfp$samesex),
  morekids = binary_summary(ae98_lfp$morekids),
  workedm  = binary_summary(ae98_lfp$workedm)
)

core_lfp <- data.frame(
  Sample = "LFP sample",
  Variable = rownames(core_lfp),
  as.data.frame(core_lfp),
  row.names = NULL
)

core_income <- rbind(
  samesex  = binary_summary(ae98_income$samesex),
  morekids = binary_summary(ae98_income$morekids),
  workedm  = binary_summary(ae98_income$workedm)
)

core_income <- data.frame(
  Sample = "Positive-income sample",
  Variable = rownames(core_income),
  as.data.frame(core_income),
  row.names = NULL
)

core_design_table <- rbind(core_lfp, core_income)

show_desc_table(
  core_design_table,
  caption = "Core design variables: instrument, treatment, and LFP outcome"
)


# ------------------------------------------------------------------------------
# 11. Outcome coding map
# ------------------------------------------------------------------------------

outcome_coding_table <- data.frame(
  Outcome_group = c(
    "Labor force participation",
    "Labor force participation",
    "Labor force participation",
    "Log income",
    "Log income",
    "Log income",
    "Log income"
  ),
  Object_name = c(
    "workedm",
    "workedm_21",
    "workedm_01",
    "lincomem_cnt",
    "lincomem",
    "lincomem_ths",
    "lincomem_hths"
  ),
  Meaning = c(
    "1 if worked, 0 otherwise",
    "2 if worked, 1 otherwise = workedm + 1",
    "1 if did not work, 0 otherwise = 1 - workedm",
    "log income measured in cents = log(100 * incomem)",
    "log income measured in dollars = log(incomem)",
    "log income measured in thousands = log(incomem / 1000)",
    "log income measured in hundred-thousands = log(incomem / 100000)"
  ),
  Sample_used = c(
    "ae98_lfp",
    "ae98_lfp",
    "ae98_lfp",
    "ae98_income",
    "ae98_income",
    "ae98_income",
    "ae98_income"
  ),
  stringsAsFactors = FALSE
)

show_desc_table(
  outcome_coding_table,
  caption = "Outcome codings used later for translation and scale sensitivity"
)


# ------------------------------------------------------------------------------
# 12. Basic continuous-variable overview
# ------------------------------------------------------------------------------

continuous_lfp <- rbind(
  agem    = continuous_summary(ae98_lfp$agem),
  agefstm = continuous_summary(ae98_lfp$agefstm)
)

continuous_lfp <- data.frame(
  Sample = "LFP sample",
  Variable = rownames(continuous_lfp),
  as.data.frame(continuous_lfp),
  row.names = NULL
)

income_summary <- rbind(
  incomem       = continuous_summary(ae98_income$incomem),
  lincomem_cnt  = continuous_summary(ae98_income$lincomem_cnt),
  lincomem      = continuous_summary(ae98_income$lincomem),
  lincomem_ths  = continuous_summary(ae98_income$lincomem_ths),
  lincomem_hths = continuous_summary(ae98_income$lincomem_hths)
)

income_summary <- data.frame(
  Sample = "Positive-income sample",
  Variable = rownames(income_summary),
  as.data.frame(income_summary),
  row.names = NULL
)

show_desc_table(
  continuous_lfp,
  caption = "Continuous covariates in the LFP sample"
)

show_desc_table(
  income_summary,
  caption = "Income variable and log-income unit transformations"
)


# ------------------------------------------------------------------------------
# 13. Final interpretation notes
# ------------------------------------------------------------------------------

cat("\nInterpretation notes:\n")
cat("  1. ae98_lfp is the sample for all labor-force-participation outcomes.\n")
cat("  2. ae98_income is smaller because log income requires incomem > 0.\n")
cat("  3. samesex should be close to 0.5 if the sex-composition instrument is plausibly random.\n")
cat("  4. The four log-income outcomes differ only by additive constants.\n")
cat("     Therefore normalized estimators should be invariant across them,\n")
cat("     while unnormalized kappa estimators may change strongly.\n")
cat("  5. X_lfp and X_income are now ready for propensity-score estimation later.\n")
cat("============================================================\n")
```


```{r}
# ==============================================================================
# DESCRIPTIVE BLOCK 2 — First stage and compliance-relevant design diagnostics
# ==============================================================================
# Requires objects from Block 1:
#   ae98_lfp, ae98_income, X_lfp, X_income
#   show_desc_table(), binary_summary()
#
# Purpose:
#   Inspect the core IV design before estimating the LATE:
#     (i)   Is Z = samesex close to randomly assigned?
#     (ii)  Does Z shift D = morekids?
#     (iii) How large is the reduced-form relationship with outcomes?
#     (iv) What is the implied simple Wald estimate before covariate adjustment?
#
# Interpretation:
#   The first stage is E[D | Z = 1] - E[D | Z = 0].
#   In this design, it should be positive: same-sex first two children increase
#   the probability of having a third child.
# ==============================================================================


cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 2 — First stage and compliance diagnostics\n")
cat("============================================================\n\n")


# ------------------------------------------------------------------------------
# 1. Helper functions
# ------------------------------------------------------------------------------

mean_by_z <- function(df, var, z = "samesex") {
  y <- df[[var]]
  zz <- df[[z]]

  c(
    Mean_Z0 = mean(y[zz == 0], na.rm = TRUE),
    Mean_Z1 = mean(y[zz == 1], na.rm = TRUE),
    Difference_Z1_minus_Z0 = mean(y[zz == 1], na.rm = TRUE) -
      mean(y[zz == 0], na.rm = TRUE),
    N_Z0 = sum(zz == 0 & !is.na(y)),
    N_Z1 = sum(zz == 1 & !is.na(y))
  )
}

simple_wald <- function(df, y_var, d_var = "morekids", z_var = "samesex") {
  y <- df[[y_var]]
  d <- df[[d_var]]
  z <- df[[z_var]]

  rf <- mean(y[z == 1], na.rm = TRUE) - mean(y[z == 0], na.rm = TRUE)
  fs <- mean(d[z == 1], na.rm = TRUE) - mean(d[z == 0], na.rm = TRUE)

  c(
    Reduced_form = rf,
    First_stage = fs,
    Wald = rf / fs
  )
}

first_stage_lm <- function(df) {
  fit <- lm(morekids ~ samesex, data = df)
  s <- summary(fit)

  data.frame(
    Term = "samesex",
    Estimate = coef(fit)["samesex"],
    Std_Error = s$coefficients["samesex", "Std. Error"],
    t_value = s$coefficients["samesex", "t value"],
    p_value = s$coefficients["samesex", "Pr(>|t|)"],
    N = nobs(fit),
    stringsAsFactors = FALSE
  )
}


# ------------------------------------------------------------------------------
# 2. Instrument balance: how many observations have Z = 0 and Z = 1?
# ------------------------------------------------------------------------------

z_distribution <- rbind(
  data.frame(
    Sample = "LFP sample",
    Variable = "samesex",
    as.data.frame(t(binary_summary(ae98_lfp$samesex))),
    row.names = NULL
  ),
  data.frame(
    Sample = "Positive-income sample",
    Variable = "samesex",
    as.data.frame(t(binary_summary(ae98_income$samesex))),
    row.names = NULL
  )
)

show_desc_table(
  z_distribution,
  caption = "Distribution of the instrument samesex"
)

cat("\nInterpretation:\n")
cat("  samesex should be close to 0.5. This is a first descriptive check that\n")
cat("  the sex-composition instrument behaves like a quasi-random assignment.\n")
cat("  Very large deviations from 0.5 would be suspicious.\n\n")


# ------------------------------------------------------------------------------
# 3. Treatment rates by instrument status: raw first stage
# ------------------------------------------------------------------------------

first_stage_table <- rbind(
  data.frame(
    Sample = "LFP sample",
    Outcome = "morekids",
    as.data.frame(t(mean_by_z(ae98_lfp, "morekids"))),
    row.names = NULL
  ),
  data.frame(
    Sample = "Positive-income sample",
    Outcome = "morekids",
    as.data.frame(t(mean_by_z(ae98_income, "morekids"))),
    row.names = NULL
  )
)

show_desc_table(
  first_stage_table,
  caption = "Raw first stage: treatment rates by samesex"
)

cat("\nInterpretation:\n")
cat("  Difference_Z1_minus_Z0 is the raw first stage.\n")
cat("  It should be positive: mothers whose first two children have the same sex\n")
cat("  are more likely to have a third child.\n")
cat("  This difference is also an empirical proxy for the complier share in the\n")
cat("  unconditional Wald design.\n\n")


# ------------------------------------------------------------------------------
# 4. Regression first stage: morekids on samesex
# ------------------------------------------------------------------------------

first_stage_reg <- rbind(
  cbind(Sample = "LFP sample", first_stage_lm(ae98_lfp)),
  cbind(Sample = "Positive-income sample", first_stage_lm(ae98_income))
)

show_desc_table(
  first_stage_reg,
  caption = "Regression first stage: lm(morekids ~ samesex)"
)

cat("\nInterpretation:\n")
cat("  The coefficient on samesex is numerically the same as the raw first-stage\n")
cat("  difference in treatment rates. The t-statistic should be very large because\n")
cat("  the AE sample is huge. Economically, however, the first stage is modest:\n")
cat("  the instrument shifts the probability of morekids by only a few percentage points.\n\n")


# ------------------------------------------------------------------------------
# 5. Outcome means by instrument status
# ------------------------------------------------------------------------------

lfp_outcome_by_z <- rbind(
  workedm = mean_by_z(ae98_lfp, "workedm"),
  workedm_21 = mean_by_z(ae98_lfp, "workedm_21"),
  workedm_01 = mean_by_z(ae98_lfp, "workedm_01")
)

lfp_outcome_by_z <- data.frame(
  Sample = "LFP sample",
  Outcome = rownames(lfp_outcome_by_z),
  as.data.frame(lfp_outcome_by_z),
  row.names = NULL
)

income_outcome_by_z <- rbind(
  lincomem_cnt = mean_by_z(ae98_income, "lincomem_cnt"),
  lincomem = mean_by_z(ae98_income, "lincomem"),
  lincomem_ths = mean_by_z(ae98_income, "lincomem_ths"),
  lincomem_hths = mean_by_z(ae98_income, "lincomem_hths")
)

income_outcome_by_z <- data.frame(
  Sample = "Positive-income sample",
  Outcome = rownames(income_outcome_by_z),
  as.data.frame(income_outcome_by_z),
  row.names = NULL
)

show_desc_table(
  lfp_outcome_by_z,
  caption = "LFP outcomes by samesex status"
)

show_desc_table(
  income_outcome_by_z,
  caption = "Log-income outcomes by samesex status"
)

cat("\nInterpretation:\n")
cat("  These are reduced-form outcome differences by instrument status.\n")
cat("  The LFP recodings should mechanically differ by sign or level, but the\n")
cat("  causal content is the same. The log-income unit transformations should have\n")
cat("  identical reduced-form differences because changing units adds a constant\n")
cat("  to log income.\n\n")


# ------------------------------------------------------------------------------
# 6. Simple unconditional Wald estimates
# ------------------------------------------------------------------------------

wald_lfp <- rbind(
  workedm = simple_wald(ae98_lfp, "workedm"),
  workedm_21 = simple_wald(ae98_lfp, "workedm_21"),
  workedm_01_raw = simple_wald(ae98_lfp, "workedm_01")
)

wald_lfp <- data.frame(
  Sample = "LFP sample",
  Outcome = rownames(wald_lfp),
  as.data.frame(wald_lfp),
  row.names = NULL
)

# For comparability with SUW reporting, add sign-flipped did-not-work version
wald_lfp$Wald_reported <- wald_lfp$Wald
wald_lfp$Wald_reported[wald_lfp$Outcome == "workedm_01_raw"] <-
  -wald_lfp$Wald[wald_lfp$Outcome == "workedm_01_raw"]

wald_income <- rbind(
  lincomem_cnt = simple_wald(ae98_income, "lincomem_cnt"),
  lincomem = simple_wald(ae98_income, "lincomem"),
  lincomem_ths = simple_wald(ae98_income, "lincomem_ths"),
  lincomem_hths = simple_wald(ae98_income, "lincomem_hths")
)

wald_income <- data.frame(
  Sample = "Positive-income sample",
  Outcome = rownames(wald_income),
  as.data.frame(wald_income),
  row.names = NULL
)

show_desc_table(
  wald_lfp,
  caption = "Simple unconditional Wald estimates: LFP outcomes"
)

show_desc_table(
  wald_income,
  caption = "Simple unconditional Wald estimates: log-income outcomes"
)

cat("\nInterpretation:\n")
cat("  The simple Wald estimate is reduced form divided by first stage.\n")
cat("  For workedm and workedm_21, the Wald estimate should be the same because\n")
cat("  adding 1 to the outcome leaves the reduced-form difference unchanged.\n")
cat("  For workedm_01, the raw sign is reversed, so Wald_reported flips it back\n")
cat("  to the SUW-style reporting convention.\n")
cat("  For log income, all four Wald estimates should be numerically identical\n")
cat("  because the log-unit transformations differ only by additive constants.\n\n")


# ------------------------------------------------------------------------------
# 7. Compact design summary for later reference
# ------------------------------------------------------------------------------

design_summary <- data.frame(
  Sample = c("LFP sample", "Positive-income sample"),
  N = c(nrow(ae98_lfp), nrow(ae98_income)),
  Share_samesex = c(
    mean(ae98_lfp$samesex, na.rm = TRUE),
    mean(ae98_income$samesex, na.rm = TRUE)
  ),
  First_stage = c(
    simple_wald(ae98_lfp, "workedm")["First_stage"],
    simple_wald(ae98_income, "lincomem")["First_stage"]
  ),
  Wald_workedm_or_lincomem = c(
    simple_wald(ae98_lfp, "workedm")["Wald"],
    simple_wald(ae98_income, "lincomem")["Wald"]
  ),
  stringsAsFactors = FALSE
)

show_desc_table(
  design_summary,
  caption = "Compact IV design summary"
)

cat("\nFinal notes for Block 2:\n")
cat("  1. If Share_samesex is close to 0.5, the instrument distribution looks plausible.\n")
cat("  2. If First_stage is positive and non-trivial, the design has empirical relevance.\n")
cat("  3. The simple Wald estimates are not the final SUW kappa estimates, but they\n")
cat("     give an intuitive first look at the direction and scale of the design.\n")
cat("  4. In later blocks, we move from this raw Wald design to covariate-adjusted\n")
cat("     propensity-score and kappa-weight diagnostics.\n")
cat("============================================================\n")

```


```{r}

# ==============================================================================
# DESCRIPTIVE BLOCK 3 — Covariate balance by instrument status
# ==============================================================================
# Requires objects from Block 1:
#   ae98_lfp, ae98_income, X_lfp, X_income
#   show_desc_table()
#
# Purpose:
#   Check whether observed covariates are balanced across Z = samesex.
#   Since samesex should be close to quasi-random, covariate differences between
#   Z = 1 and Z = 0 should be small.
#
# Important distinction:
#   This is NOT the final outcome-weight balance diagnostic.
#   This is only the raw descriptive balance of covariates by instrument status.
#   Later, outcome weights / kappa weights ask a different question:
#     Which observations drive the LATE estimate?
# ==============================================================================


cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 3 — Covariate balance by instrument status\n")
cat("============================================================\n\n")


# ------------------------------------------------------------------------------
# 1. Helper functions
# ------------------------------------------------------------------------------

smd_binary_or_continuous <- function(x, z) {
  x <- as.numeric(x)
  z <- as.numeric(z)

  x0 <- x[z == 0]
  x1 <- x[z == 1]

  m0 <- mean(x0, na.rm = TRUE)
  m1 <- mean(x1, na.rm = TRUE)

  sd_pooled <- sqrt(
    (stats::var(x0, na.rm = TRUE) + stats::var(x1, na.rm = TRUE)) / 2
  )

  smd <- ifelse(sd_pooled > 0, (m1 - m0) / sd_pooled, NA_real_)

  c(
    Mean_Z0 = m0,
    Mean_Z1 = m1,
    Diff_Z1_minus_Z0 = m1 - m0,
    SMD = smd,
    Abs_SMD = abs(smd),
    N_Z0 = sum(!is.na(x0)),
    N_Z1 = sum(!is.na(x1))
  )
}

covariate_balance_by_z <- function(df, covariates, sample_label) {
  out <- lapply(covariates, function(v) {
    stats <- smd_binary_or_continuous(df[[v]], df$samesex)

    data.frame(
      Sample = sample_label,
      Covariate = v,
      Mean_Z0 = stats["Mean_Z0"],
      Mean_Z1 = stats["Mean_Z1"],
      Diff_Z1_minus_Z0 = stats["Diff_Z1_minus_Z0"],
      SMD = stats["SMD"],
      Abs_SMD = stats["Abs_SMD"],
      N_Z0 = stats["N_Z0"],
      N_Z1 = stats["N_Z1"],
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, out)
}

flag_balance <- function(abs_smd) {
  ifelse(
    is.na(abs_smd),
    NA_character_,
    ifelse(abs_smd < 0.05, "Very small",
           ifelse(abs_smd < 0.10, "Small / acceptable", "Potential imbalance"))
  )
}


# ------------------------------------------------------------------------------
# 2. Define covariates used in the childbearing application
# ------------------------------------------------------------------------------

ae_covariates <- c(
  "agem",
  "agefstm",
  "sexk",
  "sex2nd",
  "blackm",
  "hispm",
  "othracem"
)

missing_covariates_lfp <- setdiff(ae_covariates, names(ae98_lfp))
missing_covariates_income <- setdiff(ae_covariates, names(ae98_income))

if (length(missing_covariates_lfp) > 0 || length(missing_covariates_income) > 0) {
  stop(
    paste0(
      "Some covariates are missing.\n",
      "Missing in ae98_lfp: ",
      paste(missing_covariates_lfp, collapse = ", "),
      "\nMissing in ae98_income: ",
      paste(missing_covariates_income, collapse = ", ")
    )
  )
}


# ------------------------------------------------------------------------------
# 3. Balance table: LFP sample
# ------------------------------------------------------------------------------

balance_lfp <- covariate_balance_by_z(
  df = ae98_lfp,
  covariates = ae_covariates,
  sample_label = "LFP sample"
)

balance_lfp$Balance_flag <- flag_balance(balance_lfp$Abs_SMD)

show_desc_table(
  balance_lfp,
  caption = "Raw covariate balance by samesex: LFP sample"
)

cat("\nInterpretation — LFP sample:\n")
cat("  This table compares observed covariates between Z = 0 and Z = 1.\n")
cat("  Because samesex is based on child sex composition, large differences in\n")
cat("  mother's age, age at first birth, or race variables would be suspicious.\n")
cat("  In a plausible quasi-random design, Abs_SMD values should be very small.\n\n")


# ------------------------------------------------------------------------------
# 4. Balance table: positive-income sample
# ------------------------------------------------------------------------------

balance_income <- covariate_balance_by_z(
  df = ae98_income,
  covariates = ae_covariates,
  sample_label = "Positive-income sample"
)

balance_income$Balance_flag <- flag_balance(balance_income$Abs_SMD)

show_desc_table(
  balance_income,
  caption = "Raw covariate balance by samesex: positive-income sample"
)

cat("\nInterpretation — positive-income sample:\n")
cat("  This repeats the same balance check after restricting to mothers with\n")
cat("  positive income. The income sample is selected on the outcome side, so it is\n")
cat("  useful to verify that the instrument still looks balanced in this smaller sample.\n\n")


# ------------------------------------------------------------------------------
# 5. Compact comparison across samples
# ------------------------------------------------------------------------------

balance_compact <- rbind(
  data.frame(
    Sample = "LFP sample",
    Max_Abs_SMD = max(balance_lfp$Abs_SMD, na.rm = TRUE),
    Mean_Abs_SMD = mean(balance_lfp$Abs_SMD, na.rm = TRUE),
    Covariate_with_max_Abs_SMD =
      balance_lfp$Covariate[which.max(balance_lfp$Abs_SMD)],
    stringsAsFactors = FALSE
  ),
  data.frame(
    Sample = "Positive-income sample",
    Max_Abs_SMD = max(balance_income$Abs_SMD, na.rm = TRUE),
    Mean_Abs_SMD = mean(balance_income$Abs_SMD, na.rm = TRUE),
    Covariate_with_max_Abs_SMD =
      balance_income$Covariate[which.max(balance_income$Abs_SMD)],
    stringsAsFactors = FALSE
  )
)

show_desc_table(
  balance_compact,
  caption = "Compact raw covariate-balance summary"
)

cat("\nInterpretation:\n")
cat("  Max_Abs_SMD gives the worst raw covariate imbalance by samesex.\n")
cat("  Mean_Abs_SMD summarizes the average imbalance across covariates.\n")
cat("  For this design, these values should be small because samesex should not\n")
cat("  systematically sort mothers by observed covariates.\n\n")


# ------------------------------------------------------------------------------
# 6. Optional simple plot: raw SMD by covariate
# ------------------------------------------------------------------------------

old_par <- par(no.readonly = TRUE)

plot(
  balance_lfp$Abs_SMD,
  seq_along(balance_lfp$Abs_SMD),
  xlim = c(0, max(c(balance_lfp$Abs_SMD, balance_income$Abs_SMD, 0.1), na.rm = TRUE)),
  yaxt = "n",
  xlab = "Absolute standardized mean difference by samesex",
  ylab = "",
  main = "Raw covariate balance by instrument status"
)

points(
  balance_income$Abs_SMD,
  seq_along(balance_income$Abs_SMD),
  pch = 17
)

axis(
  side = 2,
  at = seq_along(balance_lfp$Covariate),
  labels = balance_lfp$Covariate,
  las = 1
)

abline(v = 0.10, lty = 2)
abline(v = 0.05, lty = 3)

legend(
  "bottomright",
  legend = c("LFP sample", "Positive-income sample", "0.10 threshold", "0.05 reference"),
  pch = c(1, 17, NA, NA),
  lty = c(NA, NA, 2, 3),
  bty = "n"
)

par(old_par)

cat("\nPlot interpretation:\n")
cat("  This is a raw instrument-balance plot, not a final Love plot.\n")
cat("  Points close to zero mean that Z = samesex is balanced on that covariate.\n")
cat("  If the points are all far below 0.10, the descriptive instrument balance\n")
cat("  looks very clean.\n")


# ------------------------------------------------------------------------------
# 7. Save useful objects for later blocks
# ------------------------------------------------------------------------------

ae_balance_raw_lfp <- balance_lfp
ae_balance_raw_income <- balance_income
ae_balance_raw_compact <- balance_compact

cat("\nObjects created by Block 3:\n")
cat("  ae_balance_raw_lfp\n")
cat("  ae_balance_raw_income\n")
cat("  ae_balance_raw_compact\n")
cat("============================================================\n")


```


```{r}
# ==============================================================================
# DESCRIPTIVE BLOCK 4 — Propensity score and kappa-leverage diagnostic
# ==============================================================================
# Requires objects from Block 1:
#   ae98_lfp, ae98_income, X_lfp, X_income
#   show_desc_table()
#
# Purpose:
#   Diagnose whether the AE childbearing design has the same extreme-kappa problem
#   as the Card design.
#
# Key question:
#   Are there observations with extreme p(Z=1|X), especially high p but Z=0,D=1,
#   that dominate Abadie kappa weights?
#
# Interpretation:
#   In Card, p(Z|X) was highly dispersed because nearc4 is strongly predicted by
#   geography. In AE, samesex should be close to random, so p(Z|X) should stay
#   close to 0.5. That should prevent massive kappa leverage.
# ==============================================================================


cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 4 — Propensity score and kappa leverage\n")
cat("============================================================\n\n")


# ------------------------------------------------------------------------------
# 1. Helper functions
# ------------------------------------------------------------------------------

safe_logit <- function(eta) {
  eta <- pmin(pmax(eta, -35), 35)
  1 / (1 + exp(-eta))
}

logit_mle_desc <- function(Z, X) {
  Z <- as.numeric(Z)
  X <- as.matrix(X)

  fit <- glm.fit(
    x = X,
    y = Z,
    family = binomial(link = "logit")
  )

  beta <- coef(fit)
  beta[is.na(beta)] <- 0

  p <- as.vector(safe_logit(X %*% beta))
  p <- pmin(pmax(p, 1e-8), 1 - 1e-8)

  p
}

kappa_complier_desc <- function(Z, D, p) {
  Z <- as.numeric(Z)
  D <- as.numeric(D)
  p <- pmin(pmax(as.numeric(p), 1e-8), 1 - 1e-8)

  1 - D * (1 - Z) / (1 - p) - (1 - D) * Z / p
}

ps_summary <- function(p, sample_label) {
  data.frame(
    Sample = sample_label,
    N = length(p),
    Mean = mean(p),
    SD = sd(p),
    Min = min(p),
    Q01 = unname(quantile(p, 0.01)),
    Q05 = unname(quantile(p, 0.05)),
    Q25 = unname(quantile(p, 0.25)),
    Median = median(p),
    Q75 = unname(quantile(p, 0.75)),
    Q95 = unname(quantile(p, 0.95)),
    Q99 = unname(quantile(p, 0.99)),
    Max = max(p),
    Share_below_0.30 = mean(p < 0.30),
    Share_above_0.70 = mean(p > 0.70),
    Share_below_0.10 = mean(p < 0.10),
    Share_above_0.90 = mean(p > 0.90),
    stringsAsFactors = FALSE
  )
}

kappa_summary <- function(kappa, p, Z, D, sample_label) {
  abs_k <- abs(kappa)

  data.frame(
    Sample = sample_label,
    N = length(kappa),
    Mean_kappa = mean(kappa),
    SD_kappa = sd(kappa),
    Min_kappa = min(kappa),
    Q01_kappa = unname(quantile(kappa, 0.01)),
    Median_kappa = median(kappa),
    Q99_kappa = unname(quantile(kappa, 0.99)),
    Max_kappa = max(kappa),
    Max_abs_kappa = max(abs_k),
    ESS_kappa_raw = sum(kappa)^2 / sum(kappa^2),
    Share_negative_kappa = mean(kappa < 0),
    Share_Z0_D1 = mean(Z == 0 & D == 1),
    Share_Z1_D0 = mean(Z == 1 & D == 0),
    Mean_p_Z0_D1 = mean(p[Z == 0 & D == 1], na.rm = TRUE),
    Mean_p_Z1_D0 = mean(p[Z == 1 & D == 0], na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

top_kappa_summary <- function(df, p, kappa, sample_label, top_n = c(5, 10, 25, 50, 100)) {
  Z <- as.numeric(df$samesex)
  D <- as.numeric(df$morekids)
  abs_k <- abs(kappa)
  ord <- order(abs_k, decreasing = TRUE)
  total_abs <- sum(abs_k)

  rows <- lapply(top_n, function(n_top) {
    idx <- ord[seq_len(min(n_top, length(ord)))]

    data.frame(
      Sample = sample_label,
      Top_n = length(idx),
      Share_abs_kappa = sum(abs_k[idx]) / total_abs,
      Min_kappa_top = min(kappa[idx]),
      Max_kappa_top = max(kappa[idx]),
      Mean_abs_kappa_top = mean(abs_k[idx]),
      Mean_p_top = mean(p[idx]),
      Share_Z0_D1_top = mean(Z[idx] == 0 & D[idx] == 1),
      Share_Z1_D0_top = mean(Z[idx] == 1 & D[idx] == 0),
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, rows)
}


# ------------------------------------------------------------------------------
# 2. Estimate MLE propensity scores p(Z=1|X)
# ------------------------------------------------------------------------------

Z_lfp <- as.numeric(ae98_lfp$samesex)
D_lfp <- as.numeric(ae98_lfp$morekids)

Z_income <- as.numeric(ae98_income$samesex)
D_income <- as.numeric(ae98_income$morekids)

p_lfp_ml <- logit_mle_desc(Z_lfp, X_lfp)
p_income_ml <- logit_mle_desc(Z_income, X_income)

cat("MLE propensity scores estimated for:\n")
cat("  LFP sample\n")
cat("  Positive-income sample\n\n")


# ------------------------------------------------------------------------------
# 3. Propensity-score distribution
# ------------------------------------------------------------------------------

ps_dist_table <- rbind(
  ps_summary(p_lfp_ml, "LFP sample"),
  ps_summary(p_income_ml, "Positive-income sample")
)

show_desc_table(
  ps_dist_table,
  caption = "Distribution of estimated instrument propensity scores p(Z=1|X)"
)

cat("\nInterpretation:\n")
cat("  For AE, p(Z=1|X) should stay close to 0.5 because samesex is close to random.\n")
cat("  Shares below 0.30 or above 0.70 should be tiny or zero.\n")
cat("  This is very different from Card, where geography strongly predicted nearc4.\n\n")


# ------------------------------------------------------------------------------
# 4. Kappa weights for complier moments
# ------------------------------------------------------------------------------

kappa_lfp_ml <- kappa_complier_desc(Z_lfp, D_lfp, p_lfp_ml)
kappa_income_ml <- kappa_complier_desc(Z_income, D_income, p_income_ml)

kappa_dist_table <- rbind(
  kappa_summary(
    kappa = kappa_lfp_ml,
    p = p_lfp_ml,
    Z = Z_lfp,
    D = D_lfp,
    sample_label = "LFP sample"
  ),
  kappa_summary(
    kappa = kappa_income_ml,
    p = p_income_ml,
    Z = Z_income,
    D = D_income,
    sample_label = "Positive-income sample"
  )
)

show_desc_table(
  kappa_dist_table,
  caption = "Abadie kappa diagnostic under MLE p(Z|X)"
)

cat("\nInterpretation:\n")
cat("  ESS_kappa_raw = sum(kappa)^2 / sum(kappa^2) is a simple concentration check\n")
cat("  for the complier-moment kappa weights. It is not yet the final outcome-weight\n")
cat("  ESS from tau_u, but it shows whether kappa mass is dominated by a few units.\n")
cat("  If Max_abs_kappa is close to 2 and p is near 0.5, leverage is mild.\n\n")


# ------------------------------------------------------------------------------
# 5. Top absolute-kappa concentration
# ------------------------------------------------------------------------------

top_kappa_lfp <- top_kappa_summary(
  df = ae98_lfp,
  p = p_lfp_ml,
  kappa = kappa_lfp_ml,
  sample_label = "LFP sample"
)

top_kappa_income <- top_kappa_summary(
  df = ae98_income,
  p = p_income_ml,
  kappa = kappa_income_ml,
  sample_label = "Positive-income sample"
)

top_kappa_table <- rbind(top_kappa_lfp, top_kappa_income)

show_desc_table(
  top_kappa_table,
  caption = "Top absolute-kappa concentration"
)

cat("\nInterpretation:\n")
cat("  This is the compact analogue of the Card high-p diagnostic.\n")
cat("  If the top 50 or top 100 observations explain only a small share of total\n")
cat("  absolute kappa mass, then the design does not have the severe leverage pattern\n")
cat("  found in Card. In Card, top-kappa units often had high p(Z|X), Z=0, and D=1.\n")
cat("  Here, because p(Z|X) should be near 0.5, that mechanism should be much weaker.\n\n")


# ------------------------------------------------------------------------------
# 6. Cell-level check: surprising instrument cells
# ------------------------------------------------------------------------------

cell_diag <- function(df, p, kappa, sample_label) {
  Z <- as.numeric(df$samesex)
  D <- as.numeric(df$morekids)

  cells <- list(
    "Z=0, D=0" = which(Z == 0 & D == 0),
    "Z=0, D=1" = which(Z == 0 & D == 1),
    "Z=1, D=0" = which(Z == 1 & D == 0),
    "Z=1, D=1" = which(Z == 1 & D == 1)
  )

  rows <- lapply(names(cells), function(cell_name) {
    idx <- cells[[cell_name]]

    data.frame(
      Sample = sample_label,
      Cell = cell_name,
      N = length(idx),
      Share_sample = length(idx) / length(Z),
      Mean_p = mean(p[idx], na.rm = TRUE),
      Mean_kappa = mean(kappa[idx], na.rm = TRUE),
      Mean_abs_kappa = mean(abs(kappa[idx]), na.rm = TRUE),
      Max_abs_kappa = max(abs(kappa[idx]), na.rm = TRUE),
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, rows)
}

cell_diag_table <- rbind(
  cell_diag(ae98_lfp, p_lfp_ml, kappa_lfp_ml, "LFP sample"),
  cell_diag(ae98_income, p_income_ml, kappa_income_ml, "Positive-income sample")
)

show_desc_table(
  cell_diag_table,
  caption = "Kappa leverage by instrument-treatment cell"
)

cat("\nInterpretation:\n")
cat("  The dangerous Card-style cell is typically high p(Z=1|X), observed Z=0,\n")
cat("  and D=1, because 1 / (1-p) can become large. In AE, p should be close to 0.5,\n")
cat("  so even Z=0,D=1 observations should not generate huge negative kappa values.\n\n")


# ------------------------------------------------------------------------------
# 7. Save objects for later blocks
# ------------------------------------------------------------------------------

ae_p_lfp_ml <- p_lfp_ml
ae_p_income_ml <- p_income_ml

ae_kappa_lfp_ml <- kappa_lfp_ml
ae_kappa_income_ml <- kappa_income_ml

ae_ps_dist_table <- ps_dist_table
ae_kappa_dist_table <- kappa_dist_table
ae_top_kappa_table <- top_kappa_table
ae_cell_diag_table <- cell_diag_table

cat("Objects created by Block 4:\n")
cat("  ae_p_lfp_ml\n")
cat("  ae_p_income_ml\n")
cat("  ae_kappa_lfp_ml\n")
cat("  ae_kappa_income_ml\n")
cat("  ae_ps_dist_table\n")
cat("  ae_kappa_dist_table\n")
cat("  ae_top_kappa_table\n")
cat("  ae_cell_diag_table\n")
cat("============================================================\n")


```


The question when it comes to the P(Z=1|X) is can mother age, age at first birth, child-sex indicators, race, ethnicity, etc. predict whether the first two children are same-sex?
The output shows that the answer is no. As the MLE propensity is tightly distributed around 0.5
Covariates hence do not meaningfully predict samesex. This is perfect as it shows how a quasy random assignment should look alike. 
I cannot predict much better than a coin flip wheter the mother has same sex first two children



```{r}
# ==============================================================================
# DESCRIPTIVE BLOCK 5 — MLE vs CBPS propensity-score comparison
# ==============================================================================
# Requires objects from previous blocks:
#   ae98_lfp, ae98_income, X_lfp, X_income
#   ae_p_lfp_ml, ae_p_income_ml
#   show_desc_table()
#
# Purpose:
#   Compare MLE and covariate-balancing propensity scores for p(Z=1|X).
#   In the AE childbearing design, both should be close because samesex is already
#   nearly randomized with respect to X.
# ==============================================================================


cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 5 — MLE vs CBPS propensity-score comparison\n")
cat("============================================================\n\n")


# ------------------------------------------------------------------------------
# 1. CBPS-style propensity score estimator
# ------------------------------------------------------------------------------
# This solves a covariate-balancing objective:
#   sum_i Z_i X_i / p_i  ≈  sum_i (1-Z_i) X_i / (1-p_i)
#
# It is intentionally implemented as a compact diagnostic version.
# For final estimation, keep using your main replication's CBPS routine.

cbps_desc <- function(Z, X, start_p = NULL, maxit = 1000) {
  Z <- as.numeric(Z)
  X <- as.matrix(X)

  if (is.null(start_p)) {
    fit_start <- glm.fit(
      x = X,
      y = Z,
      family = binomial(link = "logit")
    )
    beta_start <- coef(fit_start)
    beta_start[is.na(beta_start)] <- 0
  } else {
    beta_start <- rep(0, ncol(X))
  }

  objective <- function(beta) {
    eta <- as.vector(X %*% beta)
    eta <- pmin(pmax(eta, -35), 35)
    p <- 1 / (1 + exp(-eta))
    p <- pmin(pmax(p, 1e-6), 1 - 1e-6)

    moments <- colMeans((Z / p - (1 - Z) / (1 - p)) * X)
    sum(moments^2)
  }

  opt <- optim(
    par = beta_start,
    fn = objective,
    method = "BFGS",
    control = list(maxit = maxit, reltol = 1e-10)
  )

  eta <- as.vector(X %*% opt$par)
  eta <- pmin(pmax(eta, -35), 35)
  p <- 1 / (1 + exp(-eta))
  p <- pmin(pmax(p, 1e-8), 1 - 1e-8)

  list(
    p = p,
    beta = opt$par,
    convergence = opt$convergence,
    objective = opt$value
  )
}


# ------------------------------------------------------------------------------
# 2. Estimate CBPS propensity scores
# ------------------------------------------------------------------------------

cbps_lfp <- cbps_desc(
  Z = ae98_lfp$samesex,
  X = X_lfp
)

cbps_income <- cbps_desc(
  Z = ae98_income$samesex,
  X = X_income
)

ae_p_lfp_cb <- cbps_lfp$p
ae_p_income_cb <- cbps_income$p

cbps_convergence <- data.frame(
  Sample = c("LFP sample", "Positive-income sample"),
  Convergence_code = c(cbps_lfp$convergence, cbps_income$convergence),
  Objective_value = c(cbps_lfp$objective, cbps_income$objective),
  stringsAsFactors = FALSE
)

show_desc_table(
  cbps_convergence,
  caption = "CBPS optimizer diagnostics"
)

cat("\nInterpretation:\n")
cat("  Convergence code 0 means successful BFGS convergence.\n")
cat("  The objective value measures remaining covariate imbalance in the CBPS moments.\n\n")


# ------------------------------------------------------------------------------
# 3. Compare MLE and CBPS propensity-score distributions
# ------------------------------------------------------------------------------

ps_summary_method <- function(p, sample_label, method_label) {
  data.frame(
    Sample = sample_label,
    Method = method_label,
    N = length(p),
    Mean = mean(p),
    SD = sd(p),
    Min = min(p),
    Q01 = unname(quantile(p, 0.01)),
    Q05 = unname(quantile(p, 0.05)),
    Q25 = unname(quantile(p, 0.25)),
    Median = median(p),
    Q75 = unname(quantile(p, 0.75)),
    Q95 = unname(quantile(p, 0.95)),
    Q99 = unname(quantile(p, 0.99)),
    Max = max(p),
    Share_below_0.30 = mean(p < 0.30),
    Share_above_0.70 = mean(p > 0.70),
    stringsAsFactors = FALSE
  )
}

ps_mle_cb_table <- rbind(
  ps_summary_method(ae_p_lfp_ml, "LFP sample", "MLE logit"),
  ps_summary_method(ae_p_lfp_cb, "LFP sample", "CBPS"),
  ps_summary_method(ae_p_income_ml, "Positive-income sample", "MLE logit"),
  ps_summary_method(ae_p_income_cb, "Positive-income sample", "CBPS")
)

show_desc_table(
  ps_mle_cb_table,
  caption = "MLE versus CBPS propensity-score distributions"
)

cat("\nInterpretation:\n")
cat("  If MLE and CBPS are very similar, this means there was little covariate\n")
cat("  imbalance for CBPS to correct. That is what we expect in the same-sex design.\n\n")


# ------------------------------------------------------------------------------
# 4. Direct MLE-CBPS difference
# ------------------------------------------------------------------------------

ps_diff_table <- data.frame(
  Sample = c("LFP sample", "Positive-income sample"),
  Mean_abs_difference = c(
    mean(abs(ae_p_lfp_cb - ae_p_lfp_ml)),
    mean(abs(ae_p_income_cb - ae_p_income_ml))
  ),
  Max_abs_difference = c(
    max(abs(ae_p_lfp_cb - ae_p_lfp_ml)),
    max(abs(ae_p_income_cb - ae_p_income_ml))
  ),
  Correlation = c(
    cor(ae_p_lfp_cb, ae_p_lfp_ml),
    cor(ae_p_income_cb, ae_p_income_ml)
  ),
  stringsAsFactors = FALSE
)

show_desc_table(
  ps_diff_table,
  caption = "Direct difference between MLE and CBPS propensity scores"
)

cat("\nInterpretation:\n")
cat("  Small mean and max differences imply that the propensity-score method does\n")
cat("  not materially change the estimated assignment probabilities.\n")
cat("  High correlation means MLE and CBPS rank observations similarly.\n\n")


# ------------------------------------------------------------------------------
# 5. Kappa leverage under CBPS
# ------------------------------------------------------------------------------

kappa_from_p <- function(Z, D, p) {
  Z <- as.numeric(Z)
  D <- as.numeric(D)
  p <- pmin(pmax(as.numeric(p), 1e-8), 1 - 1e-8)

  1 - D * (1 - Z) / (1 - p) - (1 - D) * Z / p
}

kappa_summary_short <- function(kappa, p, sample_label, method_label) {
  data.frame(
    Sample = sample_label,
    Method = method_label,
    Mean_p = mean(p),
    SD_p = sd(p),
    Min_p = min(p),
    Max_p = max(p),
    Mean_kappa = mean(kappa),
    Min_kappa = min(kappa),
    Max_kappa = max(kappa),
    Max_abs_kappa = max(abs(kappa)),
    ESS_kappa_raw = sum(kappa)^2 / sum(kappa^2),
    Share_negative_kappa = mean(kappa < 0),
    stringsAsFactors = FALSE
  )
}

ae_kappa_lfp_cb <- kappa_from_p(
  Z = ae98_lfp$samesex,
  D = ae98_lfp$morekids,
  p = ae_p_lfp_cb
)

ae_kappa_income_cb <- kappa_from_p(
  Z = ae98_income$samesex,
  D = ae98_income$morekids,
  p = ae_p_income_cb
)

kappa_mle_cb_table <- rbind(
  kappa_summary_short(ae_kappa_lfp_ml, ae_p_lfp_ml, "LFP sample", "MLE logit"),
  kappa_summary_short(ae_kappa_lfp_cb, ae_p_lfp_cb, "LFP sample", "CBPS"),
  kappa_summary_short(ae_kappa_income_ml, ae_p_income_ml, "Positive-income sample", "MLE logit"),
  kappa_summary_short(ae_kappa_income_cb, ae_p_income_cb, "Positive-income sample", "CBPS")
)

show_desc_table(
  kappa_mle_cb_table,
  caption = "Kappa leverage: MLE versus CBPS propensity scores"
)

cat("\nInterpretation:\n")
cat("  If Max_abs_kappa and ESS_kappa_raw barely change from MLE to CBPS, then\n")
cat("  propensity-score estimation is not the driver of the childbearing results.\n")
cat("  This supports the interpretation that the same-sex design is much less exposed\n")
cat("  to extreme propensity-score leverage than the Card design.\n\n")


# ------------------------------------------------------------------------------
# 6. Save objects for later blocks
# ------------------------------------------------------------------------------

ae_p_lfp_cb <- ae_p_lfp_cb
ae_p_income_cb <- ae_p_income_cb

ae_kappa_lfp_cb <- ae_kappa_lfp_cb
ae_kappa_income_cb <- ae_kappa_income_cb

ae_ps_mle_cb_table <- ps_mle_cb_table
ae_ps_diff_table <- ps_diff_table
ae_kappa_mle_cb_table <- kappa_mle_cb_table

cat("Objects created by Block 5:\n")
cat("  ae_p_lfp_cb\n")
cat("  ae_p_income_cb\n")
cat("  ae_kappa_lfp_cb\n")
cat("  ae_kappa_income_cb\n")
cat("  ae_ps_mle_cb_table\n")
cat("  ae_ps_diff_table\n")
cat("  ae_kappa_mle_cb_table\n")
cat("============================================================\n")

```






```{r}
# ==============================================================================
# DESCRIPTIVE BLOCK 6 — Outcome descriptives by treatment and instrument status
# ==============================================================================
# Requires objects from previous blocks:
#   ae98_lfp, ae98_income
#   show_desc_table()
#
# Purpose:
#   Give a concise descriptive view of the outcome side:
#     (i)   How do outcomes differ by actual treatment D = morekids?
#     (ii)  How do outcomes differ by instrument Z = samesex?
#     (iii) Are treatment-status differences much larger than instrument-status
#           differences?
#
# Interpretation:
#   Differences by D are descriptive associations and are endogenous.
#   Differences by Z are reduced-form differences and are closer to the IV logic.
# ==============================================================================


cat("\n============================================================\n")
cat("DESCRIPTIVE BLOCK 6 — Outcome descriptives by D and Z\n")
cat("============================================================\n\n")


# ------------------------------------------------------------------------------
# 1. Helper functions
# ------------------------------------------------------------------------------

group_mean_diff <- function(df, y_var, group_var) {
  y <- as.numeric(df[[y_var]])
  g <- as.numeric(df[[group_var]])

  y0 <- y[g == 0]
  y1 <- y[g == 1]

  data.frame(
    Outcome = y_var,
    Group_var = group_var,
    Mean_0 = mean(y0, na.rm = TRUE),
    Mean_1 = mean(y1, na.rm = TRUE),
    Difference_1_minus_0 = mean(y1, na.rm = TRUE) - mean(y0, na.rm = TRUE),
    SD_0 = sd(y0, na.rm = TRUE),
    SD_1 = sd(y1, na.rm = TRUE),
    N_0 = sum(!is.na(y0)),
    N_1 = sum(!is.na(y1)),
    stringsAsFactors = FALSE
  )
}

make_outcome_panel <- function(df, y_vars, sample_label, group_var) {
  out <- do.call(
    rbind,
    lapply(y_vars, function(y) group_mean_diff(df, y, group_var))
  )

  out$Sample <- sample_label
  out <- out[, c(
    "Sample",
    "Outcome",
    "Group_var",
    "Mean_0",
    "Mean_1",
    "Difference_1_minus_0",
    "SD_0",
    "SD_1",
    "N_0",
    "N_1"
  )]

  rownames(out) <- NULL
  out
}


# ------------------------------------------------------------------------------
# 2. Define outcome groups
# ------------------------------------------------------------------------------

lfp_outcomes <- c(
  "workedm",
  "workedm_21",
  "workedm_01"
)

income_outcomes <- c(
  "lincomem_cnt",
  "lincomem",
  "lincomem_ths",
  "lincomem_hths"
)


# ------------------------------------------------------------------------------
# 3. Outcomes by treatment status D = morekids
# ------------------------------------------------------------------------------

lfp_by_D <- make_outcome_panel(
  df = ae98_lfp,
  y_vars = lfp_outcomes,
  sample_label = "LFP sample",
  group_var = "morekids"
)

income_by_D <- make_outcome_panel(
  df = ae98_income,
  y_vars = income_outcomes,
  sample_label = "Positive-income sample",
  group_var = "morekids"
)

show_desc_table(
  lfp_by_D,
  caption = "LFP outcomes by treatment status D = morekids"
)

show_desc_table(
  income_by_D,
  caption = "Log-income outcomes by treatment status D = morekids"
)

cat("\nInterpretation — outcomes by D:\n")
cat("  These are raw differences between mothers with and without a third child.\n")
cat("  They are NOT causal because morekids is endogenous.\n")
cat("  They are still useful because they show the strong selection pattern behind D:\n")
cat("  mothers with more children can differ systematically in labor supply and income.\n\n")


# ------------------------------------------------------------------------------
# 4. Outcomes by instrument status Z = samesex
# ------------------------------------------------------------------------------

lfp_by_Z <- make_outcome_panel(
  df = ae98_lfp,
  y_vars = lfp_outcomes,
  sample_label = "LFP sample",
  group_var = "samesex"
)

income_by_Z <- make_outcome_panel(
  df = ae98_income,
  y_vars = income_outcomes,
  sample_label = "Positive-income sample",
  group_var = "samesex"
)

show_desc_table(
  lfp_by_Z,
  caption = "LFP outcomes by instrument status Z = samesex"
)

show_desc_table(
  income_by_Z,
  caption = "Log-income outcomes by instrument status Z = samesex"
)

cat("\nInterpretation — outcomes by Z:\n")
cat("  These are reduced-form differences by instrument status.\n")
cat("  They should be much smaller than the raw D differences because samesex only\n")
cat("  shifts treatment probability modestly. Dividing these reduced forms by the\n")
cat("  first stage gives the simple Wald estimates from Block 2.\n\n")


# ------------------------------------------------------------------------------
# 5. Compact comparison: D differences versus Z differences
# ------------------------------------------------------------------------------

compact_outcome_comparison <- data.frame(
  Sample = c(
    "LFP sample",
    "Positive-income sample"
  ),
  Main_outcome = c(
    "workedm",
    "lincomem"
  ),
  Diff_by_D = c(
    lfp_by_D$Difference_1_minus_0[lfp_by_D$Outcome == "workedm"],
    income_by_D$Difference_1_minus_0[income_by_D$Outcome == "lincomem"]
  ),
  Diff_by_Z = c(
    lfp_by_Z$Difference_1_minus_0[lfp_by_Z$Outcome == "workedm"],
    income_by_Z$Difference_1_minus_0[income_by_Z$Outcome == "lincomem"]
  ),
  First_stage = c(
    mean(ae98_lfp$morekids[ae98_lfp$samesex == 1], na.rm = TRUE) -
      mean(ae98_lfp$morekids[ae98_lfp$samesex == 0], na.rm = TRUE),
    mean(ae98_income$morekids[ae98_income$samesex == 1], na.rm = TRUE) -
      mean(ae98_income$morekids[ae98_income$samesex == 0], na.rm = TRUE)
  ),
  stringsAsFactors = FALSE
)

compact_outcome_comparison$Simple_Wald <- with(
  compact_outcome_comparison,
  Diff_by_Z / First_stage
)

compact_outcome_comparison$Abs_Diff_by_D_over_Abs_Diff_by_Z <- with(
  compact_outcome_comparison,
  abs(Diff_by_D) / abs(Diff_by_Z)
)

show_desc_table(
  compact_outcome_comparison,
  caption = "Compact comparison: treatment associations versus reduced forms"
)

cat("\nInterpretation:\n")
cat("  Diff_by_D is the raw association with actual fertility behavior.\n")
cat("  Diff_by_Z is the reduced-form effect of the instrument.\n")
cat("  The reduced form should be much smaller because the instrument only moves\n")
cat("  treatment for compliers, not for the whole sample.\n")
cat("  Simple_Wald rescales the reduced form by the first stage.\n\n")


# ------------------------------------------------------------------------------
# 6. Mechanical checks for outcome recodings
# ------------------------------------------------------------------------------

mechanical_checks <- data.frame(
  Check = c(
    "workedm_21 - workedm",
    "workedm + workedm_01",
    "lincomem_cnt - lincomem",
    "lincomem - lincomem_ths",
    "lincomem - lincomem_hths"
  ),
  Expected_value = c(
    1,
    1,
    log(100),
    log(1000),
    log(100000)
  ),
  Observed_mean = c(
    mean(ae98_lfp$workedm_21 - ae98_lfp$workedm, na.rm = TRUE),
    mean(ae98_lfp$workedm + ae98_lfp$workedm_01, na.rm = TRUE),
    mean(ae98_income$lincomem_cnt - ae98_income$lincomem, na.rm = TRUE),
    mean(ae98_income$lincomem - ae98_income$lincomem_ths, na.rm = TRUE),
    mean(ae98_income$lincomem - ae98_income$lincomem_hths, na.rm = TRUE)
  ),
  stringsAsFactors = FALSE
)

mechanical_checks$Abs_error <- abs(
  mechanical_checks$Observed_mean - mechanical_checks$Expected_value
)

show_desc_table(
  mechanical_checks,
  caption = "Mechanical checks for outcome recodings"
)

cat("\nInterpretation:\n")
cat("  These checks confirm that the alternative outcome codings are exactly\n")
cat("  translations/sign transformations of the same underlying outcome.\n")
cat("  Therefore differences across normalized and unnormalized estimators later\n")
cat("  cannot be attributed to different data; they come from estimator properties.\n\n")


# ------------------------------------------------------------------------------
# 7. Save objects
# ------------------------------------------------------------------------------

ae_lfp_by_D <- lfp_by_D
ae_income_by_D <- income_by_D
ae_lfp_by_Z <- lfp_by_Z
ae_income_by_Z <- income_by_Z
ae_outcome_compact <- compact_outcome_comparison
ae_outcome_mechanical_checks <- mechanical_checks

cat("Objects created by Block 6:\n")
cat("  ae_lfp_by_D\n")
cat("  ae_income_by_D\n")
cat("  ae_lfp_by_Z\n")
cat("  ae_income_by_Z\n")
cat("  ae_outcome_compact\n")
cat("  ae_outcome_mechanical_checks\n")
cat("============================================================\n")
```


So mothers with at least three children are about 12.1 percentage points less likely to work and have about 21.5 log points lower income than mothers with only two children.
But this is not the causal effect. It is just the raw association. Mothers who have a third child may differ systematically from mothers who do not: preferences, household structure, education, labor-market attachment, family background, partner income, etc.



The instrument on the other hand produces much smaller outcome differences 
So mothers whose first two children have the same sex are about 0.8 percentage points less likely to work and have about 0.9 log points lower income.


The first stage rescales the reduced form into the wald estimate:
This says: samesex increases the probability of having a third child by about 6.0 percentage points in the LFP sample and 5.4 percentage points in the positive-income sample.
So Wald of -0.132: Among compliers — mothers whose fertility is shifted by the same-sex composition of the first two children — having an additional child is associated with around a 13 percentage point reduction in labor-force participation in the simple unconditional Wald design.
For income : Among compliers in the positive-income sample, the simple unconditional Wald estimate suggests a negative income effect of about 16 log points.


This means the raw fertility association is about 15 times larger than the instrument reduced form for LFP, and about 24 times larger for income.

That is not a problem. It shows two things:

Treatment is endogenous/selected. Mothers with more children differ a lot from mothers without more children.
The instrument is subtle. samesex only nudges fertility behavior for a small complier group.

This is exactly why the IV estimate divides the small reduced form by the small first stage.

This is the raw treatment difference and why it is 15 times larger :
Raw treatment association = -0.121
Reduced form by instrument = -0.008
abs(-0.121) / abs(-0.008) ≈ 15.4